### 2021-09-23 01--prepare_reference_database.Rmd
make site public

---
title: "Reference database preparation"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

`dada2` requires a special format for assigning taxonomy using a reference database. 
Also, we are using two different loci and amplifying both fungi and oomycetes, so we will need a reference database for each locus.
The databases should have a format like:

```
>Kingom;Phylum;Class;Order;Family;Genus;Species;Reference
ACGAATGTGAAGTAA......
```

Where, "Reference" is a database-specific ID for each sequence, since there can be multiple sequences for a single species.
I will start the taxonomy at the Domain of Eukaryota, since it is shared with fungi.

## Preparation

### Parameters

```{r}
seed <- 1
set.seed(seed)
```

### Packages used

```{r message=FALSE}
library(dada2)
library(metacoder)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(taxize)
library(dplyr)
library(DT)
library(ape)
```

### Output location

Any data generated by this analysis that is used in other analyses is stored in `intermediate_data`.
The formatted reference database FASTA files will be stored here.

```{r}
formatted_ref_dir <- file.path("intermediate_data", "reference_databases")
if (!dir.exists(formatted_ref_dir)) {
  dir.create(formatted_ref_dir, recursive = TRUE)
}
```

## Rps10 database

### Make FASTA headers

The rps10 database described by the associated publication will be used to assign taxonomy to ASVs/OTUs for the the rps10 MiSeq sequences.
Any unidentified sequences will be blasted to genbank for identification when analyzing non-target amplification, so no non-target sequences will be added here.
The sequences for species used in the mock community are already included in this database.

```{r}
rps10_db <- read_fasta(file.path("raw_data", "reference_databases", "2020-11-3_release_2_rps10.fasta"))
rps10_data <- str_match(names(rps10_db), pattern = "name=(.+)\\|strain=(.+)\\|ncbi_acc=(.+)\\|ncbi_taxid=(.+)\\|oodb_id=(.+)\\|taxonomy=(.+)$")
colnames(rps10_data) <- c("header", "name", "strain", "ncbi_acc", "ncbi_taxid", "oodb_id", "taxonomy")
rps10_data <- as_tibble(rps10_data)
datatable(rps10_data)
```

Lets remove the "cellular_organisms" root:

```{r}
rps10_data$taxonomy <- gsub(rps10_data$taxonomy, pattern = 'cellular_organisms;', replacement = '', fixed = TRUE)
head(rps10_data$taxonomy)
```

and add a "Heterokontophyta" rank, so it has the same number of ranks as the ITS database (see below):

```{r}
rps10_data$taxonomy <- gsub(rps10_data$taxonomy, pattern = 'Eukaryota;', replacement = 'Eukaryota;Heterokontophyta;', fixed = TRUE)
head(rps10_data$taxonomy)
```

This taxonomy has the genus and species joined together as a single level.
I will split them into their own levels:

```{r}
binomial <- map_chr(str_split(rps10_data$taxonomy, pattern = ';'), `[`, 7)
genus <- map_chr(str_split(binomial, pattern = '_'), `[`, 1)
unique(genus)
rps10_data$taxonomy <- map_chr(seq_along(rps10_data$taxonomy), function(index) {
  sub(rps10_data$taxonomy[index], pattern = binomial[index], replacement = paste0(genus[index], ';', binomial[index]))
})
head(rps10_data$taxonomy)
```

I will also add a "rank" for the name of the database and reference sequence, using the index as ID:

```{r}
rps10_data$taxonomy <- paste0(rps10_data$taxonomy, ';', 'oodb_', seq_along(rps10_data$taxonomy))
head(rps10_data$taxonomy)
```

Finally, add a trailing `;` to conform to the dada2 examples:

```{r}
rps10_data$taxonomy <- paste0(rps10_data$taxonomy, ';')
head(rps10_data$taxonomy)
```

and remove any white space:

```{r}
rps10_data$taxonomy <- trimws(rps10_data$taxonomy)
rps10_db <- trimws(rps10_db)
```

Lets check that the taxonomy is now formatted as expected:

```{r}
stopifnot(all(str_count(rps10_data$taxonomy, pattern = ";") == 9))
```

We can then associate these classifications with the sequences they were derived from as save it as a new fasta file:

```{r}
rps10_ref_path <- file.path(formatted_ref_dir, "rps10_reference_db.fa")
paste0(">", rps10_data$taxonomy, "\n", rps10_db) %>%
  write_lines(file = rps10_ref_path)
```

Lets make a table of the number of sequences in each genus:

```{r}
genus_count <- table(map_chr(strsplit(rps10_data$name, split = '_'), `[`, 1))
count_table <- as.data.frame(genus_count, stringsAsFactors = FALSE)
count_table <- as_tibble(count_table)
names(count_table) <- c('Genus', 'Number of sequences')
count_table
write_csv(count_table, file = file.path("results", "rps10_genus_count_table.csv"))
```

## ITS1 database

I will use a database which is a combination of the sequences from @robideau2011dna, PhytophthoraDB, and UNITE.
I will also add some sequences for the mock community that we sequenced to make sure they are included.

### Robideau 2011

These are sequences from a study that did a phylogeny of all oomycetes using ITS and COX.

```{r}
rob_2011_seqs <- read_fasta(file.path('raw_data', 'reference_databases', 'robideau_2011_its_database.fa'))
rob_2011 <- extract_tax_data(names(rob_2011_seqs), regex = "(.*)", key = "class", class_sep = "|") %>%
  filter_taxa(!is_leaf) %>%
  filter_taxa(! grepl(taxon_names, pattern = '^unclassified_'))
stopifnot(all(rob_2011$data$tax_data$input == names(rob_2011_seqs)))
rob_2011$data$seqs <- setNames(rob_2011_seqs, rob_2011$data$tax_data$taxon_id)
rob_2011$data <- rob_2011$data["seqs"]
rob_2011
```

### UNITE

UNITE contains fungal species and some oomycetes.

```{r}
unite <- parse_unite_general(file = file.path('raw_data', 'reference_databases', 'sh_general_release_dynamic_all_02.02.2019.fasta'))
unite
```

### Phytophthora DB

Phytophthora DB contains phytophthora sequences.

```{r}
phyto_db_seqs <- read_fasta(file.path('raw_data', 'reference_databases', 'phytophthora_db_its_database.fa'))
base_class <- lookup_tax_data('Phytophthora', type = 'taxon_name') %>% classifications() %>% tail(1)
phyto_db_class <- paste0(base_class, ';', str_match(names(phyto_db_seqs), '.+ \\((.+)\\)')[, 2])
phyto_db <- extract_tax_data(phyto_db_class, regex = "(.*)", key = "class", class_sep = ";") %>%
  filter_taxa(taxon_names != 'cellular organisms')
phyto_db$data$seqs <- setNames(phyto_db_seqs, phyto_db$data$tax_data$taxon_id)
phyto_db$data <- phyto_db$data["seqs"]
phyto_db
```

### Mock community sequences

Not all the species used in the mock community have ITS1 sequences in the databases used so far. Here are counts of the number of sequences for each mock community member:

```{r}
mc_data <- read_csv(file.path('raw_data', 'mock_community.csv'))
# mc_data$species <- gsub(mc_data$species, pattern = ' ', replacement = '_')
vapply(gsub(mc_data$species, pattern = ' ', replacement = '_'), FUN.VALUE = numeric(1), function(s) {
  sum(agrepl(c(names(rob_2011_seqs), names(phyto_db_seqs), unite$data$tax_data$organism), pattern = s))
})
```

I therefore included Sanger sequences of the mock community members produced for this study.

```{r}
its_mc_seqs <- read_fasta(file.path('raw_data', 'reference_databases', 'mock_comm_its1_sanger.fasta'))
names(its_mc_seqs)
```

This file does not have their full taxonomy, so I will use the taxonomy from the rps10 database for the same species:

```{r}
names(its_mc_seqs) <- str_match(names(its_mc_seqs), pattern = '([A-z]+ [A-z]+) ?.*$')[, 2]
names(its_mc_seqs) <- gsub(names(its_mc_seqs), pattern = ' ', replacement = '_')
names(its_mc_seqs) <- map_chr(names(its_mc_seqs), function(n) {
  tax <- rps10_data$taxonomy[grep(rps10_data$name, pattern = n)]
  str_match(tax, pattern = paste0('^(.+', n, ').+$'))[1, 2] # match up until species name
})
head(names(its_mc_seqs))
```

### Make header taxonomy ranks consistent

The Phytophthora DB and Robideau sequences have a similar taxonomy

```{r}
rob_2011_seqs <- setNames(rob_2011$data$seqs, classifications(rob_2011)[names(rob_2011$data$seqs)])
head(names(rob_2011_seqs))
phyto_db_seqs <- setNames(phyto_db$data$seqs, classifications(phyto_db)[names(phyto_db$data$seqs)])
head(names(phyto_db_seqs))
```

But Unite is different:

```{r}
unite_seqs <- setNames(unite$data$tax_data$unite_seq, classifications(unite)[unite$data$tax_data$taxon_id])
head(names(unite_seqs))
```

The Phytophthora DB and robideau sequences have a these ranks:

> Domain, Kingdom, Class, Order, Family, Genus, Species

But Unite, there is also Phylum:

> Kingdom, Phylum, Class, Order, Family, Genus, Species

In Unite, Stramenopiles is called Stramenopila and is considered a Kingdom.

I can make the two the same, by adding the Eukaryota Domain to Unite and the Phylum Heterokontophyta to the other two, making the ranks:

> Domain, Kingdom, Phylum, Class, Order, Family, Genus, Species

```{r}
names(unite_seqs) <- paste0('Eukaryota;', names(unite_seqs))
names(rob_2011_seqs) <- sub(names(rob_2011_seqs), pattern = ';Stramenopiles;',
                            replacement = ';Heterokontophyta;Stramenopiles;', fixed = TRUE)
names(phyto_db_seqs) <- sub(names(phyto_db_seqs), pattern = ';Stramenopiles;',
                            replacement = ';Heterokontophyta;Stramenopiles;', fixed = TRUE)
names(phyto_db_seqs) <- sub(names(phyto_db_seqs), pattern = ';Sar;',
                            replacement = ';', fixed = TRUE)
```

The taxonomy of the sequences from our lab have already been put into this format.

The taxonomy now has the same number of ranks, but there are some differences in the classification of oomycetes in UNITE vs the other databases:

```{r}
head(names(unite_seqs)[grepl(names(unite_seqs), pattern = 'Oomy')])
head(names(rob_2011_seqs))
head(names(phyto_db_seqs))
head(names(its_mc_seqs))
```

So I will change them to be all the same as much as possible:

```{r}
names(unite_seqs) <- sub(names(unite_seqs), fixed = TRUE,
                         pattern = 'Stramenopila;Oomycota',
                         replacement = 'Heterokontophyta;Stramenopiles')
names(phyto_db_seqs) <- sub(names(phyto_db_seqs), fixed = TRUE,
                         pattern = 'Oomycota',
                         replacement = 'Oomycetes')
```

Now they should be the same:

```{r}
head(names(unite_seqs)[grepl(names(unite_seqs), pattern = 'Oomy')])
head(names(rob_2011_seqs))
head(names(phyto_db_seqs))
head(names(its_mc_seqs))
head(rps10_data$taxonomy)
```

### Combine sequences

I will add a "rank" for the reference too, with the name of the database the sequence came from and the source database index:

```{r}
names(unite_seqs) <- paste0(names(unite_seqs), ';', 'unite_', seq_along(unite_seqs))
names(rob_2011_seqs) <- paste0(names(rob_2011_seqs), ';', 'rob2011_', seq_along(rob_2011_seqs))
names(phyto_db_seqs) <- paste0(names(phyto_db_seqs), ';', 'phytodb_', seq_along(phyto_db_seqs))
names(its_mc_seqs) <- paste0(names(its_mc_seqs), ';', 'mock_', seq_along(its_mc_seqs))
its1_seqs <- c(unite_seqs, rob_2011_seqs, phyto_db_seqs, its_mc_seqs)
```

Note that the ID is an index of the sequence in the particular version of the database used.
Since none of the sequences were filtered out before this was assigned, this can be used to identify the particular sequence in a particular database.

### Remove unidentified sequences

Some of the databases (UNITE I think) has unidentified sequences.
I will remove these, so that they are not used when assigning ASVs and prevent a less good match to a more informative sequence:

```{r}
its1_seqs <- its1_seqs[! grepl(names(its1_seqs), pattern = 'unidentified')]
```

### Clean up formatting

Add combined index:

```{r}
names(its1_seqs) <- paste0(names(its1_seqs), '_', seq_along(its1_seqs))
head(names(its1_seqs))
```

This is the index of the sequence in the combined reference datbase.
No filtering should be done to the combined database after this.
Replace spaces with underscores:

```{r}
names(its1_seqs) <- gsub(names(its1_seqs), pattern = " ", replacement = "_")
```

Make sure they end in a `;`, like the example databases for dada2

```{r}
no_ending_semicolon <- ! endsWith(names(its1_seqs), ';')
names(its1_seqs)[no_ending_semicolon] <- paste0(names(its1_seqs)[no_ending_semicolon], ";")
```

Make sure everything is uppercase for consistently:

```{r}
its1_seqs <- toupper(its1_seqs)
```

and that there is not extra whitespace in header or sequence:

```{r}
its1_seqs <- trimws(its1_seqs)
names(its1_seqs) <- trimws(names(its1_seqs))
```

I will check for any sequences with ambiguity codes:

```{r}
sum(! grepl(its1_seqs, pattern = '^[AGCTagct]*$'))
```

Check for problems

```{r}
stopifnot(all(str_count(names(its1_seqs), pattern = ";") == 9))
stopifnot(all(nchar(its1_seqs) > 100))
stopifnot(! any(grepl(names(its1_seqs), pattern = 'NA')))
```

And finally save the results in a FASTA file

```{r}
its_ref_path <- file.path(formatted_ref_dir, "its1_reference_db.fa")
paste0(">", names(its1_seqs), "\n", its1_seqs) %>%
  write_lines(its_ref_path)
```

## Mock community sequences

### Look up synonyms for mock community

Many of the mock community members have synonyms and since multiple databases are used, multiple names might be used for the same taxa.
Therefore, I will look up the synonyms of taxa used in the mock community from COL.
The code below used to work, but the current version of `taxize` [has stopped supporting COL](https://github.com/ropensci/taxize/issues/796):

```{r eval=FALSE}
mc_data <- read_csv(file.path('raw_data', 'mock_community.csv'))
mc_syn_data <- as_tibble(synonyms_df(synonyms(mc_data$species, db = "col")))
mc_syn_data <- rename(mc_syn_data,  mc_name = .id, col_id = id, syn_name = name)
write_csv(mc_syn_data, file = file.path('intermediate_data', 'mock_comm_synonyms.csv'))
```

So I took the result of that code when it did work and load it here:

```{r}
mc_syn_data <- read_csv(file.path('raw_data', 'mock_comm_synonyms.csv'))
datatable(mc_syn_data)
```

### Check that mock community sequences are in reference databases

In order to fairly evaluate the performance of each primer pair / locus, we need to know if the mock community sequences are actually present in the reference databases.

```{r}
its_seqs <- read_fasta(its_ref_path)
rps10_seqs <- read_fasta(rps10_ref_path)
```

I will use `agrep` instead of `grep` to allow for any misspellings.

```{r}
is_in_db <- function(species, db, allow_ambig = TRUE) {
  names(db) <- gsub(names(db), pattern = '_', replacement = ' ')
  mc_syn_data <- readr::read_csv(file.path('raw_data', 'mock_comm_synonyms.csv'), col_types = cols())
  out <- purrr::map_lgl(species, function(s) {
    sp_names <- unique(c(s, mc_syn_data$syn_name[mc_syn_data$mc_name == s]))
    sp_names <- gsub(sp_names, pattern = '_', replacement = ' ')
    if (allow_ambig) {
      out <- any(vapply(sp_names, FUN.VALUE = logical(1), function(name) {
        any(grepl(names(db), pattern = name, ignore.case = TRUE))
      }))
    } else {
      out <- any(vapply(sp_names, FUN.VALUE = logical(1), function(name) {
        any(grepl(names(db), pattern = name, ignore.case = TRUE) & grepl(db, pattern = '^[AGCTagct]*$'))
      }))
    }
    return(out)
  })
  names(out) <- species
  out
}

mc_data$in_its_db <- is_in_db(mc_data$species, its_seqs)
mc_data$in_rps10_db <- is_in_db(mc_data$species, rps10_seqs)
```

Some of the methods used to assign taxonomy cannot use sequences with ambiguity codes, so I will check for sequences with no ambiguity codes.

```{r}
mc_data$in_its_db_no_ambig <- is_in_db(mc_data$species, its_seqs, allow_ambig = FALSE)
mc_data$in_rps10_db_no_ambig <- is_in_db(mc_data$species, rps10_seqs, allow_ambig = FALSE)
```

I will save a version of the mock community data that has these results included for other analyses

```{r}
write_csv(mc_data, file = file.path('intermediate_data', 'mock_community.csv'))
datatable(mc_data)
```

All the mock community species have sequences for both rps10 and ITS1 in the reference database:

```{r}
stopifnot(all(mc_data$in_its_db & mc_data$in_rps10_db))
```

### Make phylogeny of mock community sequences

This might be useful for seeing which species cant be distinguished.
Note that this is the phylogeny of the whole reference sequence, not just the amplicon.

```{r fig.width=10, fig.height=10}
nj_tree <- function(seqs, ...) {
  # Align sequences:
  aligned <- seqs %>%
    insect::char2dna() %>%
    ips::mafft(method = 'localpair', exec = '/usr/bin/mafft')

  # Make distance matrix
  dist <- ape::dist.dna(aligned, ...)

  # Make tree
  tree <- ape::nj(dist)
  tree <- ape::ladderize(tree)
  tree <- phangorn::midpoint(tree)

  tree
}

is_in_mc <- function(db) {
  mc_sp <- unique(c(mc_data$species, mc_syn_data$syn_name))
  names(db) <- gsub(names(db), pattern = '_', replacement = ' ')
  grepl(names(db), pattern = paste0(mc_sp, collapse = '|'))
}

plot_mc_tree <- function(db) {
  seqs <- db[is_in_mc(db)]
  db_names <- str_match(names(seqs), pattern = '.+;(.+);.+;$')[, 2]
  db_names <- gsub(db_names, pattern = '_', replacement = ' ')
  mc_names <- mc_syn_data$mc_name[match(db_names, mc_syn_data$syn_name)]
  names(seqs) <- db_names
  has_syn <- ! is.na(mc_names) & mc_names != db_names
  names(seqs)[has_syn] <- paste0(names(seqs[has_syn]), ' (', mc_names[has_syn], ')')
  seqs <- seqs[! duplicated(names(seqs))]
  tree <- nj_tree(seqs, model = 'N')
  plot.phylo(tree)
  axis(side = 1)
  title(xlab = 'Number of differing sites')
}

plot_mc_tree(rps10_seqs)
plot_mc_tree(its_seqs)
```

## Software used

```{r}
sessioninfo::session_info()
```

## References

### 2021-09-23 02--preparation_and_quality_filtering.Rmd
make site public

---
title: "Read processing and quality filtering"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

This is an analysis of the data from the MiSeq run testing the rps10 barcode and associated primers.
Multiple mock community samples and environmental samples were sequenced.
This will roughly follow the [DADA2 ITS Pipeline Workflow (1.8)](https://benjjneb.github.io/dada2/ITS_workflow.html) and the [DADA2 Pipeline Tutorial (1.12)](https://benjjneb.github.io/dada2/tutorial.html).

## Prepare

### Notes on how to use this analysis

Some of the long running operations that produce output files only run if their output does not exist.
To rerun them, delete the corresponding file in the intermediate data folder.

### Packages used

```{r message=FALSE}
library(dada2)
library(ShortRead)
library(Biostrings)
library(dplyr)
library(purrr)
library(furrr)
library(tidyr)
library(readr)
library(ggplot2)
library(gridExtra)
library(sessioninfo)
```

### Parameters

```{r}
seed <- 1
expected_error_filter_limit <- 5 # Default: no expected-error filtering (Inf)
truncation_qual_limit <- 5 # Default: 2
set.seed(seed)
```

### Parallel processing

Commands that have "future" in them are run on multiple cores using the `furrr` and `future` packages.

```{r}
plan(multiprocess)
```

## Read in raw data

### FASTQ data

```{r}
fastq_dir_path <- "raw_data/sequences"
fastq_paths <- list.files(fastq_dir_path, pattern = "\\.fastq")
fastq_data <- tibble(file_id = sub(fastq_paths, pattern = "\\.fastq\\.gz$", replacement = ""),
                     sample_id = sub(fastq_paths, pattern = "_.+$", replacement = ""),
                     direction = c("Reverse", "Forward")[grepl(fastq_paths, pattern = "R1") + 1],
                     raw_path = file.path(fastq_dir_path, fastq_paths))
print(fastq_data)
```

### Primer data

There were three different pairs of primers used.
I renamed the primers manually from "cons", "its", and "f" in Felipe's original sample metadata file.
I will make two tables: one for the individual primers and one for the pairs of primers used.

```{r}
primer_data_path <- file.path("raw_data", "primer_data.csv")
primer_data <- read_csv(primer_data_path)
print(primer_data)
```

```{r}
primer_pair_data_path <- file.path("raw_data", "primer_pair_data.csv")
primer_pair_data <- read_csv(primer_pair_data_path)
print(primer_pair_data)
```

### Sample metadata

I will edit the sample metadata some to make it more how I like it.

```{r}
if (! dir.exists("intermediate_data")) {
  dir.create("intermediate_data")
}
metadata_path <- file.path("raw_data", "metadata.csv")
modified_metadata_path <- file.path("intermediate_data", "metadata.csv")
metadata <- read_csv(metadata_path) %>%
  arrange(well) %>%
  rename(sample_id = well, primer_pair_id = primer, dna_type = origin, dna_sample_id = sample) %>%
  left_join(primer_pair_data, by = c("primer_pair_id" = "felipes_id")) %>%
  mutate(primer_pair_id = primer_pair_id.y) %>%
  select(-primer_pair_id.y)
```

I will also add a column to identify the type of sample:

```{r}
metadata$sample_type <- 'Sample'
metadata$sample_type[grepl(metadata$dna_type, pattern = 'control')] <- 'Negative control'
metadata$sample_type[grepl(metadata$dna_type, pattern = 'mock')] <- 'Mock community'
```

and I will save an edited version in the intermediate files to use in other analyses.

```{r}
write_csv(metadata, modified_metadata_path)
print(metadata)
```

## Check that forward and reverse reads are in matching order

One of the assumptions of this analysis is that forward and reverse reads are in matching order, so I will check that on one of the FASTQ files.

```{r}
paired_file_paths <- fastq_data %>%
  filter(sample_id == first(sample_id)) %>%
  pull(raw_path)
get_read_names <- function(path) {
  seqs <- readFastq(path)
  sub(as.character(seqs@id), pattern = " .+$", replacement = "")
}
stopifnot(all(get_read_names(paired_file_paths[1]) == get_read_names(paired_file_paths[2])))
```

## Trim primer sequences

Since it is possible for some small amplicons to read completely though, we have to check for both primer in both read, and just to be safe, in all possible orientations.

### Prefilter Ns

Since Ns can make it difficult to map short sequences, I will remove any sequences that have Ns and make a new set of sequence files.

```{r}
prefiltered_read_dir <- file.path("intermediate_data", "prefiltered_sequences")
fastq_data$prefiltered_path <- file.path(prefiltered_read_dir, base::basename(fastq_data$raw_path))
```

This will take a while:

```{r}
if (! all(file.exists(fastq_data$prefiltered_path))) {
  filterAndTrim(fwd = fastq_data[fastq_data$direction == "Forward", ][["raw_path"]], 
                filt = fastq_data[fastq_data$direction == "Forward", ][["prefiltered_path"]], 
                rev = fastq_data[fastq_data$direction == "Reverse", ][["raw_path"]], 
                filt.rev = fastq_data[fastq_data$direction == "Reverse", ][["prefiltered_path"]], 
                maxN = 0,
                multithread = TRUE)
}
```

### Check for primers in sequences

For each primer I will generate the reverse, complement, and reverse complement, so I can check for any unusual read orientation.
First I will need to get the sequences for each of the possible orientations.

```{r}
primer_data$complement <- map_chr(primer_data$sequence, function(x) toString(complement(DNAString(x))))
primer_data$reverse <- map_chr(primer_data$sequence, function(x) toString(reverse(DNAString(x))))
primer_data$rev_comp <- map_chr(primer_data$sequence, function(x) toString(reverseComplement(DNAString(x))))
primer_hit_data <- gather(primer_data, key = "orientation", value = "sequence", sequence, complement, reverse, rev_comp)
print(primer_hit_data)
```

This is a function copied from the [DADA2 ITS Pipeline Workflow (1.8)](https://benjjneb.github.io/dada2/ITS_workflow.html) to count number of times a sequence appears in a fastq.

```{r}
primer_hits <- function(primer, path) {
    # Counts number of reads in which the primer is found
    nhits <- vcountPattern(primer, sread(readFastq(path)), fixed = FALSE)
    return(sum(nhits > 0))
}
```

This will take a while:

```{r}
primer_hit_data_csv_path <- file.path("intermediate_data", "primer_hit_data.csv")
if (file.exists(primer_hit_data_csv_path)) {
  primer_hit_data <- read_csv(primer_hit_data_csv_path)
} else {
  primer_hit_counts <- future_map(fastq_data$prefiltered_path, 
                                  function (a_path) map_dbl(primer_hit_data$sequence, primer_hits, path = a_path))
  names(primer_hit_counts) <- fastq_data$file_id
  primer_hit_data <- bind_cols(primer_hit_data, as_tibble(primer_hit_counts))
  write_csv(primer_hit_data, primer_hit_data_csv_path)
}
print(primer_hit_data)
```

The complement and reverse of any of the primers are never detected:

```{r}
primer_hit_data %>%
  filter(orientation %in% c("complement", "reverse")) %>%
  select(!!! fastq_data$file_id) %>%
  sum()
```

So I will remove those rows to make things easier to analyze.

```{r}
primer_hit_data <- filter(primer_hit_data, ! orientation %in% c("complement", "reverse"))
```

Now I can look at what primers were detected in which samples, grouping samples by the primer pair used.

```{r}
primer_hit_data %>%
  gather(key = "file_id", value = "count", !!! fastq_data$file_id) %>%
  mutate(sample_id = sub(file_id, pattern = "_R[12]$", replacement = "")) %>%
  left_join(metadata, by = "sample_id") %>%
  left_join(fastq_data, by = "file_id") %>%
  select(primer_id, primer_direction = direction.x, orientation, count, sample_id = sample_id.x, primer_pair_id, forward, reverse, read_direction = direction.y) %>%
  ggplot(aes(x = primer_id, y = count, color = orientation)) +
  geom_boxplot() +
  facet_grid(read_direction ~ primer_pair_id)
```

It looks like the expected primer was found most often and in ITS6/7 and rps10_felipe there was some reads that extended into the other primer since the reverse complement of the other primer was found at a lower rate.
The rps10_f-F primer was also found in the samples with the standard rps10-F primers.
This is to be expected, since the rps10_f-F primers sequences occurs in the rps10_Final amplicon sequences (i.e.  rps10_Final is a bigger amplcion that contains rps10_Felipe).

If we look at just how many primer sequences appeared in samples sequenced with different primer pairs, we see a similar pattern:

```{r}
per_primer_pair_hits <- primer_hit_data %>%
  gather(key = "file_id", value = "count", !!! fastq_data$file_id) %>%
  mutate(sample_id = sub(file_id, pattern = "_R[12]$", replacement = "")) %>%
  left_join(metadata, by = "sample_id") %>%
  left_join(fastq_data, by = "file_id") %>%
  select(primer_id, primer_direction = direction.x, orientation, count, sample_id = sample_id.x, primer_pair_id, forward, reverse, read_direction = direction.y) %>%
  group_by(primer_pair_id, primer_id) %>%
  summarise(count = sum(count)) %>%
  group_by(primer_pair_id) %>%
  mutate(percent = count / sum(count) * 100)
print(per_primer_pair_hits)
```

Plotting makes it apparent that the correct primers are found the vast majority of the time:

```{r}
primer_plot <-  per_primer_pair_hits %>%
  ggplot(aes(x = primer_id, y = percent)) +
  geom_bar(stat = "identity") +
  facet_wrap(. ~ primer_pair_id)
print(primer_plot)
```

However, log scaling shows that there are some unexpected primers occurring in around 0.1% to 0.01% of reads.

```{r}
primer_plot + 
  scale_y_log10()
```

We can quickly calculate the total percentage of unexpected primer hit using the fact that they are always less than 1% of reads:

```{r}
sum(per_primer_pair_hits$count[per_primer_pair_hits$percent < 1]) / sum(per_primer_pair_hits$count)
```

This might represent the index switching known to occur in MiSeq sequencing.

### Remove primers with cutadapt

First lets check if `cutadapt` is installed.

```{r}
tryCatch(system2("cutadapt", args = "--version"),
         warning=function(w) {
           stop("cutadapt cannot be found on PATH. Is it installed?")
         })
```

We next need a place to put the results of both reads that were trimmed successfully and those that were not.

```{r}
trimmed_read_dir <- file.path("intermediate_data", "trimmed_sequences")
if (! dir.exists(trimmed_read_dir)) {
  dir.create(trimmed_read_dir)
}
fastq_data$trimmed_path <- file.path(trimmed_read_dir, paste0(fastq_data$file_id, ".fastq.gz"))

untrimmed_read_dir <- file.path("intermediate_data", "untrimmed_sequences")
if (! dir.exists(untrimmed_read_dir)) {
  dir.create(untrimmed_read_dir)
}
fastq_data$untrimmed_path <- file.path(untrimmed_read_dir, paste0(fastq_data$file_id, ".fastq.gz"))
```

Since there are three different sets of primers used, three different types of `cutadapt` commands must be applied.
The code below makes a `cutadapt` command for each sample.

```{r}
cutadapt_data <- metadata %>%
  select(sample_id, forward_id = forward, reverse_id = reverse) %>%
  left_join(primer_data[, c("primer_id", "sequence", "rev_comp")], by = c("forward_id" = "primer_id")) %>%
  rename(forward_seq = sequence, forward_seq_rc = rev_comp) %>%
  left_join(primer_data[, c("primer_id", "sequence", "rev_comp")], by = c("reverse_id" = "primer_id")) %>%
  rename(reverse_seq = sequence, reverse_seq_rc = rev_comp) %>%
  select(-forward_id, -reverse_id)

cutadapt_data <- fastq_data %>%
  filter(direction == "Forward") %>%
  select(sample_id, prefiltered_path, trimmed_path, untrimmed_path) %>%
  right_join(cutadapt_data, by = "sample_id") %>%
  rename(forward_input_path = prefiltered_path,
         forward_output_path = trimmed_path, 
         forward_untrimmed_path = untrimmed_path)

cutadapt_data <- fastq_data %>%
  filter(direction == "Reverse") %>%
  select(sample_id, prefiltered_path, trimmed_path, untrimmed_path) %>%
  right_join(cutadapt_data, by = "sample_id") %>%
  rename(reverse_input_path = prefiltered_path,
         reverse_output_path = trimmed_path,
         reverse_untrimmed_path = untrimmed_path)

cutadapt_info_dir <- file.path("intermediate_data", "cutadapt_info")
if (! dir.exists(cutadapt_info_dir)) {
  dir.create(cutadapt_info_dir)
}
cutadapt_data$info_path <- file.path(cutadapt_info_dir, paste0(cutadapt_data$sample_id, ".txt"))

cutadapt_data <- cutadapt_data %>%
  mutate(command_args = paste(
    "-g", forward_seq,
    "-a", reverse_seq_rc,
    "-G", reverse_seq,
    "-A", forward_seq_rc,
    "-n", 2, # -n 2 required to remove FWD and REV from reads
    "-o", forward_output_path,
    "-p", reverse_output_path,
    "--minimum-length", 50,
    "--untrimmed-output", forward_untrimmed_path,
    "--untrimmed-paired-output", reverse_untrimmed_path,
    "--quiet",
    # "--info-file", info_path,
    forward_input_path,
    reverse_input_path
  ))
print(cutadapt_data)
```

And now we can run the commands. **This will take a while**.

```{r cutadapt}
if (! all(file.exists(c(fastq_data$trimmed_path, fastq_data$untrimmed_path)))) {
  cutadapt_output <- future_map(cutadapt_data$command_args, ~system2("cutadapt", args = .x))
}
```

### Check that primers were removed

This will take a while:

```{r}
cutadapt_verify_path <- file.path("intermediate_data", "cutadapt_verify_data.csv")
if (file.exists(cutadapt_verify_path)) {
  cutadapt_verify_data <- read_csv(cutadapt_verify_path)
} else {
  primer_hit_counts <- future_map(fastq_data$trimmed_path, 
                                  function (a_path) map_dbl(primer_hit_data$sequence, primer_hits, path = a_path))
  names(primer_hit_counts) <- fastq_data$file_id
  cutadapt_verify_data <- primer_hit_data
  cutadapt_verify_data[names(primer_hit_counts)] <- primer_hit_counts
  write_csv(cutadapt_verify_data, cutadapt_verify_path)
}
print(cutadapt_verify_data)
```

We can now do a similar graph to what was done before:

```{r}
 cutadapt_plot <- cutadapt_verify_data %>%
  gather(key = "file_id", value = "count", !!! fastq_data$file_id) %>%
  mutate(sample_id = sub(file_id, pattern = "_R[12]$", replacement = "")) %>%
  left_join(metadata, by = "sample_id") %>%
  left_join(fastq_data, by = "file_id") %>%
  select(primer_id, primer_direction = direction.x, orientation, count, sample_id = sample_id.x, primer_pair_id, forward, reverse, read_direction = direction.y) %>%
  ggplot(aes(x = primer_id, y = count + 0.1, color = orientation)) +
  geom_boxplot() +
  facet_grid(read_direction ~ primer_pair_id)
print(cutadapt_plot)
```

The rps10_f is expected to be found in rps10_final, so that OK.
If we log scale the y axis we can see there are some other primers left over, but not many in most samples.

```{r}
cutadapt_plot +
  scale_y_log10()
```

## Check read quality

First I will make plots for each sample:

```{r message=FALSE}
sample_quality_plot_dir <- file.path("results", "sample_quality_plots")
if (! dir.exists(sample_quality_plot_dir)) {
  dir.create(sample_quality_plot_dir, recursive = TRUE)
}
sample_quality_plots <- future_map(metadata$sample_id,
                                   .options = furrr_options(seed = seed),
                                   function(id) {
                                     title <- paste0(metadata[metadata$sample_id == id, 1:3], collapse = "--") %>%
                                       gsub(pattern = "/", replacement = "_", fixed = TRUE)
                                     plot <- plotQualityProfile(fastq_data$trimmed_path[fastq_data$sample_id == id]) +
                                       ggtitle(title)
                                     ggsave(plot, 
                                            filename = paste0(title, ".pdf"), 
                                            width = 7, height = 7 / 1.618, 
                                            path = sample_quality_plot_dir)
                                     return(plot)
})
```

And plots for each primer pair:

```{r message=FALSE}
primer_quality_plots <- future_map(unique(metadata$primer_pair_id),
                                   .options = furrr_options(seed = seed),
                                   function(id) {
                                     sample_ids <- metadata$sample_id[metadata$primer_pair_id == id]
                                     forward_paths <- fastq_data %>%
                                       filter(sample_id %in% sample_ids, direction == "Forward") %>%
                                       pull(trimmed_path)
                                     reverse_paths <- fastq_data %>%
                                       filter(sample_id %in% sample_ids, direction == "Reverse") %>%
                                       pull(trimmed_path)
                                     forward_plot <- plotQualityProfile(forward_paths) +
                                       ggtitle(paste(id, "Forward"))
                                     reverse_plot <- plotQualityProfile(forward_paths) +
                                       ggtitle(paste(id, "Reverse"))
                                     combined_plot <- gridExtra::grid.arrange(forward_plot, reverse_plot)
                                     ggsave(combined_plot, 
                                            filename = paste0(gsub(id, pattern = "/", replacement = "_", fixed = TRUE), ".pdf"), 
                                            width = 10, height = 16.18, 
                                            path = sample_quality_plot_dir)
                                     plot(combined_plot)
                                     return(combined_plot)
                                   })
```

And plots for each sample type:

```{r message=FALSE}
primer_quality_plots <- future_map(unique(metadata$dna_type), 
                                   .options = furrr_options(seed = seed),
                                   function(type) {
                                     sample_ids <- metadata$sample_id[metadata$dna_type == type]
                                     forward_paths <- fastq_data %>%
                                       filter(sample_id %in% sample_ids, direction == "Forward") %>%
                                       pull(trimmed_path)
                                     reverse_paths <- fastq_data %>%
                                       filter(sample_id %in% sample_ids, direction == "Reverse") %>%
                                       pull(trimmed_path)
                                     forward_plot <- plotQualityProfile(forward_paths) +
                                       ggtitle(paste(type, "Forward"))
                                     reverse_plot <- plotQualityProfile(forward_paths) +
                                       ggtitle(paste(type, "Reverse"))
                                     combined_plot <- gridExtra::grid.arrange(forward_plot, reverse_plot)
                                     ggsave(combined_plot, 
                                            filename = paste0(type, ".pdf"), 
                                            width = 10, height = 16.18, 
                                            path = sample_quality_plot_dir)
                                     plot(combined_plot)
                                     return(combined_plot)
                                   })
```

## Quality filter reads

Now we can do the main quality filtering for the reads.
Here reads with Ns, below a length of 50, or that are expected to have more than `r expected_error_filter_limit` errors will be removed.
First I will define a place to put the results:

```{r}
filtered_reads_dir <- file.path("intermediate_data", "filtered_sequences")
fastq_data$filtered_path <- file.path(filtered_reads_dir, paste0(fastq_data$file_id, ".fastq.gz"))
print(fastq_data)
```

**This might take a while.**

```{r}
if (! dir.exists(filtered_reads_dir)) {
  filter_results <- filterAndTrim(fwd = fastq_data$trimmed_path[fastq_data$direction == "Forward"], 
                                  filt = fastq_data$filtered_path[fastq_data$direction == "Forward"],
                                  rev =  fastq_data$trimmed_path[fastq_data$direction == "Reverse"], 
                                  filt.rev = fastq_data$filtered_path[fastq_data$direction == "Reverse"], 
                                  maxN = 0, 
                                  maxEE = c(expected_error_filter_limit, expected_error_filter_limit), 
                                  truncQ = truncation_qual_limit, 
                                  minLen = 50, 
                                  rm.phix = TRUE, 
                                  compress = TRUE,
                                  multithread = TRUE)
  filter_results <- as_tibble(filter_results)
  print(colMeans(filter_results))
}
```

## Save data sets

Some of the tables used here are used in other Rmarkdown files and will be saved in CSV files.

```{r}
write_csv(fastq_data, file.path("intermediate_data", "fastq_data.csv"))
write_csv(primer_data, file.path("intermediate_data", "primer_data.csv"))
write_csv(metadata, file.path("intermediate_data", "metadata.csv"))
```

## Software information

```{r}
sessioninfo::session_info()
```

### 2021-09-23 03--abundance_matrix_preparation.Rmd
make site public

---
title: "Abundance matrix and taxonomic assignment"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

This is an analysis of the data from the MiSeq run testing the rps10 barcode and associated primers.
Multiple mock community samples and environmental samples were sequenced.
This will roughly follow the [DADA2 ITS Pipeline Workflow (1.8)](https://benjjneb.github.io/dada2/ITS_workflow.html) and the [DADA2 Pipeline Tutorial (1.12)](https://benjjneb.github.io/dada2/tutorial.html).

## Prepare

### Notes on how to use this analysis

Some of the long running operations that produce output files only run if their output does not exist.
To rerun them, delete the corresponding file in the intermediate data folder.

### Packages used

```{r message=FALSE}
library(dada2)
library(ShortRead)
library(Biostrings)
library(dplyr)
library(purrr)
library(furrr)
library(tidyr)
library(readr)
library(ggplot2)
library(gridExtra)
library(sessioninfo)
library(Biostrings)
library(stringr)
library(metacoder)
```

### Parameters

```{r}
seed <- 9999
set.seed(seed)
min_read_merging_overlap <- 15 # Default is 12
max_read_merging_mismatch <- 2 # Default is 0
remove_chimeras <- TRUE # Default is TRUE
min_asv_length <- 50
its_clustering_threshold <- 0.99
rps10_clustering_threshold <- 0.96
```

### Parallel processing

Commands that have "future" in them are run on multiple cores using the `furrr` and `future` packages.

```{r}
plan(multiprocess)
```

## Learn the Error Rates

Error rates of incorrect base calls during sequencing must be estimated to do ASV calling.
This process will estimate those error rates from the data.
First I will load the data for the fastq files for each sample that was generated previously.

```{r}
fastq_data <- read_csv(file.path("intermediate_data", "fastq_data.csv"))
```

and join this with the sample metadata so that I can distinguish rps10 from ITS1 samples.

```{r}
metadata <- read_csv(file.path('intermediate_data', 'metadata.csv'))
fastq_data <- metadata %>%
  select(sample_id, locus, primer_pair_id) %>%
  right_join(fastq_data, by = "sample_id") %>%
  mutate(file_name = paste0(file_id, '.fastq.gz'))
print(fastq_data)
```

To simplify the following code, I will make a function to get the fastq file paths for a particular combination of primer pair and read direction.

```{r}
get_fastq_paths <- function(my_direction, my_primer_pair_id) {
  fastq_data %>%
    filter(direction == my_direction, primer_pair_id == my_primer_pair_id, file.exists(filtered_path)) %>%
    pull(filtered_path)
}
```

Next, I will make a function to infer the error profile (for each type of nucleotide mutation) for each a given read direction (forward/reverse) and primer pair, and use that information to infer ASVs using dada2.

```{r}
infer_asvs <- function(my_direction, my_primer_pair_id, plot_error_rates = TRUE) {
  # Get relevant FASTQ files 
  fastq_paths <- get_fastq_paths(my_direction, my_primer_pair_id)

  # Infer error rates for each type of nucleotide mutation
  error_profile <- learnErrors(fastq_paths, multithread = TRUE)

  # Plot error rates
  if (plot_error_rates) {
    cat(paste0('Error rate plot for the ', my_direction, ' read of primer pair ', my_primer_pair_id, ' \n'))
    print(plotErrors(error_profile, nominalQ = TRUE))
  }

  # Infer ASVs
  asv_data <- dada(fastq_paths, err = error_profile, multithread = TRUE)
  return(asv_data)
}
```

Now I can infer the ASVs for each sample, with different error profiles for each combination of read direction and primer pair.

**This will take a while.**

```{r asv_inference}
denoised_data_path <- file.path("intermediate_data", "denoised_data.Rdata")
if (file.exists(denoised_data_path)) {
  load(denoised_data_path)
} else {
  run_dada <- function(direction) {
    lapply(unique(fastq_data$primer_pair_id), function(primer_pair_id) infer_asvs(direction, primer_pair_id)) %>%
      unlist(recursive = FALSE)
  }
  dada_forward <- run_dada("Forward")
  dada_reverse <- run_dada("Reverse")
  save(dada_forward, dada_reverse, file = denoised_data_path)
}
```

## Merge paired reads

This will combine the forward and reverse reads into a single read based on overlaps.

```{r read_merging, message=FALSE}
merged_read_data_path <- file.path('intermediate_data', 'merged_reads.rds')
if (file.exists(merged_read_data_path)) {
  merged_reads <- readRDS(merged_read_data_path)
} else {
  merged_reads <- mergePairs(dadaF = dada_forward,
                             derepF = file.path('intermediate_data', 'filtered_sequences', names(dada_forward)),
                             dadaR = dada_reverse,
                             derepR = file.path('intermediate_data', 'filtered_sequences', names(dada_reverse)),
                             minOverlap = min_read_merging_overlap,
                             maxMismatch = max_read_merging_mismatch,
                             returnRejects = TRUE, 
                             verbose = TRUE)
  saveRDS(merged_reads, file = merged_read_data_path)
}
```

I will plot the amount of overlap and percent identity in the overlap region to get an idea of how each locus is getting merged.
First I will combine all the read merging output into a single table with a new column for which sample it came from:

```{r}
non_empty_merged_reads <- merged_reads[map_dbl(merged_reads, nrow) > 0]
merge_data <- non_empty_merged_reads %>%
  bind_rows() %>%
  mutate(file_name = rep(names(non_empty_merged_reads), map_int(non_empty_merged_reads, nrow)),
         sample_id = gsub(file_name, pattern = '_.+$', replacement = '')) %>%
  as_tibble()
```

Next I will add columns for the metadata so I can tell which samples are for each locus

```{r}
metadata <- read_csv(file.path('intermediate_data', 'metadata.csv'))
merge_data <- left_join(merge_data, metadata, by = 'sample_id')
```

and remove any unneeded columns

```{r}
merge_data <- select(merge_data, locus, nmatch, nmismatch, nindel, accept)
merge_data
```

I will add new columns for overlap length and percent ID:

```{r}
merge_data <- mutate(merge_data,
                     overlap = nmatch + nmismatch,
                     mismatch = nmismatch + nindel,
                     identity = (overlap - mismatch) / overlap)
```

and now I can reformat the data for plotting and plot

```{r fig.width=8, fig.height=5}
merge_plot <- merge_data %>%
  select(locus, mismatch,  accept, overlap) %>%
  rename('Locus' = locus, 'Mismatches and Indels' = mismatch, 'Merged' = accept, 'Overlap Length' = overlap) %>%
  gather(key = 'stat', value = 'value', -Locus, -Merged) %>%
  ggplot(aes(x = value, fill = Merged)) +
  facet_grid(Locus ~ stat, scales = 'free') +
  geom_histogram(bins = 30) +
  scale_fill_viridis_d(begin = 0.8, end = 0.2) +
  labs(x = '', y = 'ASV count', fill = 'Merged') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position="bottom") 
ggsave(merge_plot, filename = 'read_merging.png', path = 'results', width = 8, height = 8)
merge_plot
```

## Create ASV abundance matrix

This will create the long-sought-after abundance matrix (ASV table).

```{r message=FALSE}
raw_abundance_data <- map(merged_reads, function(x) filter(x, accept == TRUE)) %>%
  makeSequenceTable()
hist(nchar(getSequences(raw_abundance_data)))
```

## Create OTU abundance matrix

I will also create an OTU abundance matrix so I can evaluate the two methods with an OTU-based approch.
Clustering is a greedy algorithm with sequences presorted by abundance and automatically masked for low-complexity regions.

```{r}
vserach_cluster <- function(seqs, seq_abund, id_threshold = 0.97, method = "fast") {
  # Check that VSEARCH is installed
  tryCatch(system2("vsearch", args = "--version", stdout = FALSE, stderr = FALSE),
           warning=function(w) {
             stop("vsearch cannot be found on PATH. Is it installed?")
           })

  # Run VSEARCH
  # seqs <- seqs[order(seq_abund, decreasing = TRUE)]
  input_fasta_path <- tempfile()
  write_lines(paste0('>', seq_along(seqs), ';size=', seq_abund, '\n', seqs), path = input_fasta_path)
  otu_centroid_path <- tempfile()
  command_args <- paste(paste0("--cluster_", method), 
                        input_fasta_path,
                        "--threads", detectCores() - 1,
                        "--id", id_threshold,
                        "--sizein",
                        "--strand plus",
                        "--fasta_width 0", # 0 = no wrapping in fasta file
                        "--centroids", otu_centroid_path)
  system2("vsearch", args = command_args, stdout = FALSE, stderr = FALSE)

  # Return OTU sequences
  centroids <- read_fasta(otu_centroid_path)
  names(centroids) <- str_match(names(centroids), pattern = 'size=(.+)$')[, 2]
  return(centroids)
}

merged_read_seqs <- unlist(map(merged_reads, function(x) {
  x$sequence[x$accept]
}))
unique_merged_read_seqs <- unique(merged_read_seqs)
length(unique_merged_read_seqs)
unique_read_counts <- map_dbl(unique_merged_read_seqs, function(s) {
  sum(map_dbl(merged_reads, function(sample_data) {
    sum(sample_data$abundance[sample_data$sequence == s & sample_data$accept])
  }))
})
otu_its_seqs <- vserach_cluster(seqs = unique_merged_read_seqs,
                               seq_abund = unique_read_counts,
                               id_threshold = its_clustering_threshold,
                               method = 'size') %>%
  toupper()
otu_rps10_seqs <- vserach_cluster(seqs = unique_merged_read_seqs,
                               seq_abund = unique_read_counts,
                               id_threshold = rps10_clustering_threshold,
                               method = 'size') %>%
  toupper()
```

Now I will create the OTU abundance matrix in the same format as dada2 outputs.

```{r}
metadata <- read_csv(file.path('intermediate_data', 'metadata.csv'))
otus_per_sample <- map(rownames(raw_abundance_data), function(sample) {
  sample_id <- str_match(sample, pattern = '^(.+)_.+$')[, 2]
  if (metadata$locus[metadata$sample_id == sample_id] == "rps10") {
    otu_seqs <- otu_rps10_seqs
  } else {
    otu_seqs <- otu_its_seqs
  }
  merged_read_data <- merged_reads[[sample]]
  sample_otu_counts <- map_int(otu_seqs, function(s) {
    sum(merged_read_data$abundance[merged_read_data$sequence == s & merged_read_data$accept])
  })
  names(sample_otu_counts) <- otu_seqs
  all_unique_otus <- unique(c(otu_rps10_seqs, otu_its_seqs))
  out <- as.integer(rep(0, length(all_unique_otus)))
  names(out) <- all_unique_otus
  out[names(sample_otu_counts)] <- sample_otu_counts
  out
  return(out)
})

raw_otu_abundance_data <- do.call(rbind, otus_per_sample)
rownames(raw_otu_abundance_data) <- rownames(raw_abundance_data)
```

and remove and OTUs with no data (these might be OTUs for rps10 clustered at the 99% level for example)

```{r}
raw_otu_abundance_data <- raw_otu_abundance_data[, colSums(raw_otu_abundance_data) > 0]
```

## Chimera removal

**This might take a while**

```{r remove_chimeras}
if (remove_chimeras) {
  # ASVs
  asv_abundance_data <- removeBimeraDenovo(raw_abundance_data,
                                           method = "consensus", 
                                           multithread = TRUE, 
                                           verbose = TRUE)
  dim(asv_abundance_data)
  print(sum(asv_abundance_data)/sum(raw_abundance_data))

  # OTUs
  otu_abundance_data <- removeBimeraDenovo(raw_otu_abundance_data,
                                           method = "consensus", 
                                           multithread = TRUE, 
                                           verbose = TRUE)
  dim(otu_abundance_data)
  print(sum(otu_abundance_data)/sum(raw_otu_abundance_data))
} else {
  asv_abundance_data <- raw_abundance_data
  otu_abundance_data <- raw_otu_abundance_data
}
```

## Remove short sequences

Sequences that are less than 50 cannot be assigned a taxonomy.

```{r}
asv_abundance_data <- asv_abundance_data[, nchar(colnames(asv_abundance_data)) >= min_asv_length]
otu_abundance_data <- otu_abundance_data[, nchar(colnames(otu_abundance_data)) >= min_asv_length]
```

## Assign taxonomy

Since there are two loci used, I will need to use two different reference databases.
First I will split abundance matrix in to RPS10 and ITS samples:

```{r}
fastq_data
rps10_abund_asv <- asv_abundance_data[rownames(asv_abundance_data) %in% fastq_data$file_name[fastq_data$locus == "rps10"], ]
its_abund_asv <- asv_abundance_data[rownames(asv_abundance_data) %in% fastq_data$file_name[fastq_data$locus == "ITS"], ]
rps10_abund_otu <- otu_abundance_data[rownames(otu_abundance_data) %in% fastq_data$file_name[fastq_data$locus == "rps10"], ]
its_abund_otu <- otu_abundance_data[rownames(otu_abundance_data) %in% fastq_data$file_name[fastq_data$locus == "ITS"], ]
```

Since there are two different loci, ASVs should either be in one locus or another but not both, so we can remove any ASVs that are not present in the two groups. If there is an ASV that is in both, I will assign it to the one with more reads.

```{r}
# ASVs
in_both <- colSums(rps10_abund_asv) != 0 & colSums(its_abund_asv) != 0
assign_to_its <- in_both & colSums(its_abund_asv) > colSums(rps10_abund_asv)
assign_to_rps <- in_both & colSums(its_abund_asv) < colSums(rps10_abund_asv)
is_rps <- (colSums(rps10_abund_asv) != 0 & colSums(its_abund_asv) == 0) | assign_to_rps
is_its <- (colSums(its_abund_asv) != 0 & colSums(rps10_abund_asv) == 0) | assign_to_its
rps10_abund_asv <- rps10_abund_asv[ , is_rps]
its_abund_asv <- its_abund_asv[ , is_its]

# OTUs
in_both <- colSums(rps10_abund_otu) != 0 & colSums(its_abund_otu) != 0
assign_to_its <- in_both & colSums(its_abund_otu) > colSums(rps10_abund_otu)
assign_to_rps <- in_both & colSums(its_abund_otu) < colSums(rps10_abund_otu)
is_rps <- (colSums(rps10_abund_otu) != 0 & colSums(its_abund_otu) == 0) | assign_to_rps
is_its <- (colSums(its_abund_otu) != 0 & colSums(rps10_abund_otu) == 0) | assign_to_its
rps10_abund_otu <- rps10_abund_otu[ , is_rps]
its_abund_otu <- its_abund_otu[ , is_its]
```

The number of ASVs left in the two groups should sum to the total number of ASVs, since there should be no overlap.

```{r}
stopifnot(ncol(rps10_abund_asv) + ncol(its_abund_asv) == ncol(asv_abundance_data))
stopifnot(ncol(rps10_abund_otu) + ncol(its_abund_otu) == ncol(otu_abundance_data))
```

Then I can assign the taxonomy on each database separately:

```{r}
# ASVs
tax_results_rps10_asv <- assignTaxonomy(rps10_abund_asv, 
                                        refFasta = file.path("intermediate_data", "reference_databases", "rps10_reference_db.fa"), 
                                        taxLevels = c("Domaine", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Reference"),
                                        minBoot = 0,
                                        tryRC = TRUE,
                                        outputBootstraps = TRUE,
                                        multithread = TRUE)
tax_results_its_asv <- assignTaxonomy(its_abund_asv, 
                                      refFasta = file.path("intermediate_data", "reference_databases", "its1_reference_db.fa"), 
                                      taxLevels = c("Domaine", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Reference"),
                                      minBoot = 0,
                                      tryRC = TRUE,
                                      outputBootstraps = TRUE,
                                      multithread = TRUE)

# OTUs
tax_results_rps10_otu <- assignTaxonomy(rps10_abund_otu, 
                                        refFasta = file.path("intermediate_data", "reference_databases", "rps10_reference_db.fa"), 
                                        taxLevels = c("Domaine", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Reference"),
                                        minBoot = 0,
                                        tryRC = TRUE,
                                        outputBootstraps = TRUE,
                                        multithread = TRUE)
tax_results_its_otu <- assignTaxonomy(its_abund_otu, 
                                      refFasta = file.path("intermediate_data", "reference_databases", "its1_reference_db.fa"), 
                                      taxLevels = c("Domaine", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Reference"),
                                      minBoot = 0,
                                      tryRC = TRUE,
                                      outputBootstraps = TRUE,
                                      multithread = TRUE)
```

## Align to reference sequence for percent ID

A high bootstrap value does not necessarily mean a good match to the reference sequence.
As long as the match is much better than any other match, the bootstrap will be high, even if the best match is not that great.
Therefore I will also align the ASV sequences to the reference sequence they were assigned to get a percent identity.

```{r}
its_seqs <- read_fasta(file.path('intermediate_data', 'reference_databases', 'its1_reference_db.fa'))
rps10_seqs <- read_fasta(file.path('intermediate_data', 'reference_databases', 'rps10_reference_db.fa'))

get_ref_seq <- function(tax_result, db) {
  ref_i <- as.integer(str_match(tax_result$tax[, 'Reference'], '^.+_([0-9]+)$')[ ,2])
  db[ref_i]
}

get_align_pid <- function(ref, asv) {
  mat <- nucleotideSubstitutionMatrix(match = 1, mismatch = -3, baseOnly = TRUE)
  align <-  pairwiseAlignment(pattern = asv, subject = ref, type = 'global-local')
  is_match <- strsplit(as.character(align@pattern), '')[[1]] == strsplit(as.character(align@subject), '')[[1]]
  sum(is_match) / length(is_match)
}

get_pids <- function(tax_result, db) {
  ref_seq <- get_ref_seq(tax_result, db)
  asv_seq <- rownames(tax_result$tax)
  future_map2_dbl(ref_seq, asv_seq, get_align_pid) * 100
}

rps10_pids_asv <- get_pids(tax_results_rps10_asv, rps10_seqs)
its_pids_asv <- get_pids(tax_results_its_asv, its_seqs)
rps10_pids_otu <- get_pids(tax_results_rps10_otu, rps10_seqs)
its_pids_otu <- get_pids(tax_results_its_otu, its_seqs)
```

Now I can add these PIDs into the taxonomy assignment results as another rank, with its percent identity to its assigned reference sequence as a level in the taxonomy.

```{r}
add_pid_to_tax <- function(tax_result, pid) {
  tax_result$tax <- cbind(tax_result$tax, ASV = rownames(tax_result$tax))   
  tax_result$boot <- cbind(tax_result$boot, ASV = pid)
  tax_result
}

tax_results_rps10_asv <- add_pid_to_tax(tax_results_rps10_asv, rps10_pids_asv)
tax_results_its_asv <- add_pid_to_tax(tax_results_its_asv, its_pids_asv)
tax_results_rps10_otu <- add_pid_to_tax(tax_results_rps10_otu, rps10_pids_otu)
tax_results_its_otu <- add_pid_to_tax(tax_results_its_otu, its_pids_otu)
```

## Make classification/bootstrap vector

I will combine the taxonomic assignments and bootstrap values for each locus into a single classification vector.
This will store all the taxonomic and bootstrap information in a single vector.

```{r}
assignTax_as_char <- function(res) {
  out <- vapply(1:nrow(res$tax), FUN.VALUE = character(1), function(i) {
    paste(res$tax[i, ],
          res$boot[i, ],
          colnames(res$tax), 
          sep = '--', collapse = ';')
  })
  names(out) <- rownames(res$tax)
  return(out)
}

seq_tax_asv <- c(assignTax_as_char(tax_results_rps10_asv), assignTax_as_char(tax_results_its_asv))
seq_tax_otu <- c(assignTax_as_char(tax_results_rps10_otu), assignTax_as_char(tax_results_its_otu))
```

Again, let make sure that there is a single taxonomic assignment for each ASV.

```{r}
stopifnot(all(names(seq_tax_asv) %in% colnames(asv_abundance_data)))
stopifnot(all(! duplicated(names(seq_tax_asv))))
stopifnot(all(names(seq_tax_otu) %in% colnames(otu_abundance_data)))
stopifnot(all(! duplicated(names(seq_tax_otu))))
```

## Reformat ASV table

I will reformat the abundance matrix to something I like more and is compatible with the `taxa` package.

```{r}
# ASVs
formatted_abund_asv <- t(asv_abundance_data)
colnames(formatted_abund_asv) <- sub(colnames(formatted_abund_asv), pattern = "_.+$", replacement = "")
formatted_abund_asv <- cbind(sequence = rownames(formatted_abund_asv), 
                             taxonomy = seq_tax_asv[rownames(formatted_abund_asv)], 
                             formatted_abund_asv)
formatted_abund_asv <- as_tibble(formatted_abund_asv)
write_csv(formatted_abund_asv, path = file.path('intermediate_data', 'abundance_asv.csv'))
print(formatted_abund_asv)

# OTUs
formatted_abund_otu <- t(otu_abundance_data)
colnames(formatted_abund_otu) <- sub(colnames(formatted_abund_otu), pattern = "_.+$", replacement = "")
formatted_abund_otu <- cbind(sequence = rownames(formatted_abund_otu), 
                             taxonomy = seq_tax_otu[rownames(formatted_abund_otu)], 
                             formatted_abund_otu)
formatted_abund_otu <- as_tibble(formatted_abund_otu)
write_csv(formatted_abund_otu, path = file.path('intermediate_data', 'abundance_otu.csv'))
print(formatted_abund_otu)
```

## Read/ASV counts throughout pipeline

I will track how many reads/ASVs were preserved at each step of the process in order to help identify any problems.

Get raw read counts for steps before read merging:

* raw reads
* prefilterd for Ns
* primers removed
* quality filtered

First I will make a table with the metadata and file names for each step combined:

```{r}
# Get file paths for just forward reads (counts are the same for both directions)
forward_fastq_data <- fastq_data %>%
  filter(direction == "Forward") %>%
  select(sample_id, raw_path, prefiltered_path, trimmed_path, untrimmed_path, filtered_path, file_name)

# Combine with metadata
count_data <- metadata %>%
  filter(primer_pair_id %in% c('rps10_Final', 'ITS6/7'), dna_type != 'mock1') %>%
  select(sample_id, locus, dna_type, sample_type) %>%
  left_join(forward_fastq_data, by = "sample_id")
```

Then count the reads in each file:

```{r}
count_reads_in_fastqgz <- function(path) {
  count <- system(paste('zcat', path, '|', 'wc', '-l'), intern = TRUE)
  as.numeric(count) / 4
}

count_data$raw_reads <- map_dbl(count_data$raw_path, count_reads_in_fastqgz)
count_data$n_filtered_reads <- map_dbl(count_data$prefiltered_path, count_reads_in_fastqgz)
count_data$trimmed_reads <- map_dbl(count_data$trimmed_path, count_reads_in_fastqgz)
count_data$qual_filtered_reads <- map_dbl(count_data$filtered_path, count_reads_in_fastqgz)

# remove columns no longer needed
count_data <- select(count_data, -prefiltered_path, -trimmed_path, -untrimmed_path, -filtered_path, -raw_path)
```

Get read counts after read merging:

```{r}
count_merged_reads <- function(read_data, merged) {
  if (is.null(read_data)) {
    return(0)
  }
  filter(read_data, accept == merged) %>%
    pull(abundance) %>%
    sum()
}

count_data$merged_reads <- map_dbl(merged_reads[count_data$file_name], count_merged_reads, merged = FALSE) + map_dbl(merged_reads[count_data$file_name], count_merged_reads, merged = TRUE)
count_data$merged_seqs <- map_dbl(merged_reads, nrow)[count_data$file_name]
count_data$filtered_merged_reads <- map_dbl(merged_reads[count_data$file_name], count_merged_reads, merged = TRUE)
count_data$filtered_merged_seqs <- map_dbl(merged_reads, function(x) sum(x$accept))[count_data$file_name]
```

Get read/ASV counts after asv inference:

```{r}
count_data$raw_asvs <- apply(raw_abundance_data, MARGIN = 1, function(x) sum(!is.na(x) & x > 0))[count_data$file_name]
count_data$raw_asv_reads <- apply(raw_abundance_data, MARGIN = 1, sum, na.rm = TRUE)[count_data$file_name]
```

Get read/ASV counts after chimera removal and short sequence filtering:

```{r}
count_data$chimera_filtered_asvs <- apply(asv_abundance_data, MARGIN = 1, function(x) sum(!is.na(x) & x > 0, na.rm = TRUE))[count_data$file_name]
count_data$chimera_filtered_reads <- apply(asv_abundance_data, MARGIN = 1, sum, na.rm = TRUE)[count_data$file_name]
```

Get read/ASV counts after low-abundance sequence filtering

```{r}
count_data$abund_filtered_asvs <- apply(asv_abundance_data, MARGIN = 1, function(x) sum(!is.na(x) & x >= 30, na.rm = TRUE))[count_data$file_name]
count_data$abund_filtered_reads <- apply(asv_abundance_data, MARGIN = 1, function(x) sum(x[x >= 30], na.rm = TRUE))[count_data$file_name]
```

Save data:

```{r}
write_csv(count_data, file = file.path('results', 'read_asv_counts.csv'))
```

Prepare data for plotting:

```{r}
plot_data <- pivot_longer(count_data, colnames(count_data)[-(1:5)], names_to = 'stat', values_to = 'count')
plot_data$type <- str_extract(plot_data$stat, pattern = '([a-z]+)$')
plot_data$type[plot_data$type == "seqs"] <- "asvs"
stage_key <- c(raw_reads = "Raw reads", 
               n_filtered_reads = "N prefiltered", 
               trimmed_reads = "Primers trimmed", 
               qual_filtered_reads = "Quality filtered",
               merged_reads = "Merged reads", 
               merged_seqs = "Merged reads", 
               filtered_merged_reads = "Filtered merged reads", 
               filtered_merged_seqs = "Filtered merged reads",
               raw_asvs = "Raw ASVs", 
               raw_asv_reads = "Raw ASVs",
               chimera_filtered_asvs = "Chimera/short filtered", 
               chimera_filtered_reads = "Chimera/short filtered",
               abund_filtered_asvs = "Abundance filtered", 
               abund_filtered_reads = "Abundance filtered")
plot_data$stage <- factor(stage_key[plot_data$stat], levels = unique(stage_key), ordered = TRUE)
```

Plot all samples:

```{r}
ggplot(plot_data, aes(x = stage, y = count, group = sample_id, color = locus)) + 
  facet_grid(type ~ ., scales = "free_y") +
  # theme_minimal() +
  expand_limits(y = 0) +
  # scale_y_continuous(trans='log10') +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  geom_line(aes(linetype = sample_type))
```

Plot just mock community samples:

```{r}
plot_data %>%
  filter(sample_type == "Mock community") %>%
  ggplot(aes(x = stage, y = count, group = sample_id, color = locus)) + 
  facet_grid(type ~ ., scales = "free_y") +
  # theme_minimal() +
  expand_limits(y = 0) +
  scale_y_continuous(trans='log10') +
  theme(axis.text.x = element_text(angle=45,hjust=1)) +
  geom_line()
```

## Software used

```{r}
sessioninfo::session_info()
```

### 2021-09-23 04--blast_classification.Rmd
make site public

---
title: "Alternative taxonomic assignment using BLAST"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

## Prepare

### Packages used

```{r message=FALSE}
library(dplyr)
library(purrr)
library(furrr)
library(tidyr)
library(readr)
library(ggplot2)
library(sessioninfo)
library(metacoder)
library(vegan)
library(viridis)
library(DT)
library(stringr)
library(qsubmitter)
library(taxize)
```

### Parameters

```{r}
cgrb <- remote_server$new(server = "shell.cgrb.oregonstate.edu", user = "fosterz", port  = 732)
files <- remote_server$new(server = "files.cgrb.oregonstate.edu", user = "fosterz", port  = 732)
remote_repository_path <- "/dfs/Grunwald_Lab/home/fosterz/repositories/rps10_barcode"
```

### Parameters

```{r}
seed <- 1
min_evalue <- '1e-3'
set.seed(seed)
```

## BLAST

A blast-based classification will be useful for the the non-target sequences detection.
It will also be useful for verifying the `dada2`-based classification.

### Make query file

I will make FASTA files of ASV and OTU sequences with header containing their row indexes in the abundance matrix.

```{r}
# ASVs
abundance_asv <- read_csv(file.path('intermediate_data', 'abundance_asv.csv'))
query_seq_path_asv <- file.path('intermediate_data', 'blast_query_asv.fa')
paste0('>asv_', 1:nrow(abundance_asv), '\n', abundance_asv$sequence) %>%
  write_lines(file = query_seq_path_asv)

# OTUs
abundance_otu <- read_csv(file.path('intermediate_data', 'abundance_otu.csv'))
query_seq_path_otu <- file.path('intermediate_data', 'blast_query_otu.fa')
paste0('>otu_', 1:nrow(abundance_otu), '\n', abundance_otu$sequence) %>%
  write_lines(file = query_seq_path_otu)
```

### Run BLAST

First I will need to transfer the file from this computer to the CGRB cluster.

```{r blast_upload_query}
# ASVs
remote_query_path_asv <- file.path(remote_repository_path, 'blast_query_asv.fa')
rsync_push(local_path = query_seq_path_asv, remote_path = remote_query_path_asv, remote = files)

# OTUs
remote_query_path_otu <- file.path(remote_repository_path, 'blast_query_otu.fa')
rsync_push(local_path = query_seq_path_otu, remote_path = remote_query_path_otu, remote = files)
```

Then I can run blast remotely.

```{r blast_run}
out_cols <- 'qseqid sallseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore staxids sskingdoms score sscinames'
run_blast_remotely <- function(query_path, out_path, ...) {
  blast_command <- paste('blastn',
                         "-query", query_path,
                         "-db nt",
                         "-dust no",
                         paste("-evalue", min_evalue),
                         # "-perc_identity 70",
                         paste0("-outfmt '10 ", out_cols, "'"),
                         "-max_hsps 1",
                         "-max_target_seqs 100",
                         '-num_threads 8',
                         '-out', out_path)
  qsub(command = blast_command,
       remote = cgrb,
       remote_cwd = remote_repository_path,
       cores = 8,
       queue = 'bpp@!(uncia)', # For some reason, I was getting errors on uncia, so this avoids that node
       ...)
}

# ASVs
remote_blast_out_asv <- file.path(remote_repository_path, 'blast_result_asv.csv')
run_blast_remotely(query_path = remote_query_path_asv, out_path = remote_blast_out_asv)

# OTUs
remote_blast_out_otu <- file.path(remote_repository_path, 'blast_result_otu.csv')
run_blast_remotely(query_path = remote_query_path_otu, out_path = remote_blast_out_otu)
```

and download the results:

```{r blast_download}
# ASVs
blast_result_path_asv <- file.path('intermediate_data', 'blast_results_asv.csv')
rsync_pull(local_path = blast_result_path_asv, remote_path = remote_blast_out_asv, remote = files)

# OTUs
blast_result_path_otu <- file.path('intermediate_data', 'blast_results_otu.csv')
rsync_pull(local_path = blast_result_path_otu, remote_path = remote_blast_out_otu, remote = files)
```

and read them into R:

```{r}
blast_results_asv <- read_csv(blast_result_path_asv, col_names = strsplit(out_cols, split = ' ')[[1]], col_types = 'ccddddddddddcccc')
blast_results_otu <- read_csv(blast_result_path_otu, col_names = strsplit(out_cols, split = ' ')[[1]], col_types = 'ccddddddddddcccc')
```

There might be some warnings about parsing errors caused by `,` in the `sscinames` column, but these should not affect the analysis since that information is not used and there are no columns after it to mess up.
I will select the best hit for each ASV based on e-value and percent identity:

```{r}
select_best_blast_hit <- function(blast_results) {
  blast_results %>%
    group_by(qseqid) %>%
    filter(evalue == min(evalue)) %>%
    filter(pident == max(pident)) %>%
    filter(row_number() == 1) # break ties by picking first value
}

blast_results_asv <- select_best_blast_hit(blast_results_asv)
blast_results_otu <- select_best_blast_hit(blast_results_otu)
```

Then I can look up the taxonomic info from the NCBI taxonomy database using the taxon ID that blast returns:

```{r}
lookup_tax <- function(blast_results) {
  blast_results$staxids <- sub(blast_results$staxids, pattern = ';.+$', replacement = '')
  classification(as.uid(unique(blast_results$staxids), check = FALSE), db = 'ncbi')
}

blast_class_asv <- lookup_tax(blast_results_asv)
blast_class_otu <- lookup_tax(blast_results_otu)
```

and add that to the results table

```{r}
get_classification <- function(x) {
  if (is.logical(x)) {
    return(NA)
  } else {
    return(paste(x$name, collapse = ';'))
  }
}

blast_results_asv$blast_tax <- map_chr(blast_class_asv, get_classification)[blast_results_asv$staxids]
blast_results_otu$blast_tax <- map_chr(blast_class_otu, get_classification)[blast_results_otu$staxids]
```

and combine that with the abundance matrix

```{r}
add_to_abund <- function(abundance, blast_results) {
  blast_results %>%
    ungroup() %>%
    transmute(sequence = abundance$sequence[as.numeric(sub(qseqid, pattern = '^asv_|otu_', replacement = ''))],
              blast_pid = pident,
              blast_cov = c(qend - qstart) / nchar(sequence) * 100,
              blast_tax = blast_tax) %>%
    right_join(abundance, by = 'sequence')
}

abundance_asv <- add_to_abund(abundance_asv, blast_results_asv)
abundance_otu <- add_to_abund(abundance_otu, blast_results_otu)
```

Finally, lets save that modified matrix for further analyses.
Note that this overwrites the abundance matrix.

```{r}
write_csv(abundance_asv, file.path('intermediate_data', 'abundance_asv.csv'))
write_csv(abundance_otu, file.path('intermediate_data', 'abundance_otu.csv'))
```

## Software used

```{r}
sessioninfo::session_info()
```

### 2021-09-23 05--otu_clustering_threshold.Rmd
make site public

---
title: "Clustering ASVs into OTUs"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

## Prepare

### Packages used

```{r message=FALSE}
library(dplyr)
library(purrr)
library(furrr)
library(tidyr)
library(readr)
library(ggplot2)
library(sessioninfo)
library(metacoder)
library(vegan)
library(viridis)
library(DT)
library(stringr)
library(qsubmitter)
library(taxize)
library(forcats)
library(parallel)
```

### Parameters

```{r}
opt_thresh_range = seq(0.9, 1, by = 0.001) # The specific thresholds checked during the optimization
opt_min_read_count <- 50 # used in id threshold optimization
opt_cluster_method <- 'size' # 'fast' for length-ordered, 'size' for abundance-ordered
seed <- 1
set.seed(seed)
```

## Read abundance matrix and sample data

```{r}
metadata <- read_csv(file.path('intermediate_data', 'metadata.csv'))
abundance <- read_csv(file.path('intermediate_data', 'abundance_asv.csv'))
print(metadata)
```

Some samples do not have sequences so I will remove those from the metadata table:

```{r}
metadata <- metadata[metadata$sample_id %in% colnames(abundance), ]
```

## Make function to run VSEARCH

I will want to check for an optimal clustering threshold for RPS10 since it is not currently known. 
To do this, I will need to cluster over many thresholds. 
It is important to note that the optimal threshold is relative to the reference sequence database and the clustering algorithm used, therefore this analysis is mostly to get a rough idea of an optimal threshold.

## Find optimal clustering threshold for mock community

First I will filter sample metadata to just mock community and the primers used in the publication.
We are using only the mock community samples since that is the closest thing to a natural community that we know the composition of.
I did something similar with the reference sequences, but they all resulted in too few clusters even if sequences were clustered at 100% (i.e. unique sequences).
On reflection, this is probably because reference databases have more diversity and fewer erroneous sequences than real data from a natural sample would.
Our environmental samples could not be used because we do not know the "correct" number of species.

```{r}
mock_meta <- filter(metadata, dna_type == 'mock2', primer_pair_id %in% c('ITS6/7', 'rps10_Final'))
mock_meta
```

I will need the number of the mock community species added later, so I will import the data for the mock community composition:

```{r}
mc_data <- read_csv(file.path('intermediate_data', 'mock_community.csv'))
```

I will also zero out low abundance ASVs, since that would be done in a normal metabarcoding analysis.
Here we are using a rather high minimum abundance of `r opt_min_read_count` since this is a low-complexity sample, and thus higher read count per species, relative to most natural communities, so erroneous sequence abundance is probably similarly inflated.

```{r}
opt_filtered_abund <- select(abundance, sequence, !!! mock_meta$sample_id)
opt_filtered_abund <- opt_filtered_abund[rowSums(opt_filtered_abund[, mock_meta$sample_id]) > opt_min_read_count, ]
opt_filtered_abund
```

Count the number of OTUs for each sample over a range of clustering thresholds.

```{r}
vserach_cluster <- function(seqs, seq_abund, id_threshold = 0.97, method = "fast") {
  # Check that VSEARCH is installed
  tryCatch(system2("vsearch", args = "--version", stdout = FALSE, stderr = FALSE),
           warning=function(w) {
             stop("vsearch cannot be found on PATH. Is it installed?")
           })

  # Run VSEARCH
  input_fasta_path <- tempfile()
  write_lines(paste0('>', seq_along(seqs), ';size=', seq_abund, '\n', seqs), path = input_fasta_path)
  otu_centroid_path <- tempfile()
  command_args <- paste(paste0("--cluster_", method), 
                        input_fasta_path,
                        "--threads", detectCores() - 1,
                        "--id", id_threshold,
                        "--sizein",
                        "--strand plus",
                        "--fasta_width 0", # 0 = no wrapping in fasta file
                        "--centroids", otu_centroid_path)
  system2("vsearch", args = command_args, stdout = FALSE, stderr = FALSE)

  # Return OTU sequences
  centroids <- read_fasta(otu_centroid_path)
  names(centroids) <- str_match(names(centroids), pattern = 'size=(.+)$')[, 2]
  return(centroids)
}

opt_thresh_data <- tibble(thresh = opt_thresh_range)
opt_thresh_data[mock_meta$locus] <- map(mock_meta$sample_id, function(sample_id) {
  map_dbl(opt_thresh_range, function(threshold) {
    has_seq <- opt_filtered_abund[[sample_id]] > 0
    length(vserach_cluster(seqs = opt_filtered_abund$sequence[has_seq],
                           seq_abund = opt_filtered_abund[[sample_id]][has_seq],
                           id_threshold = threshold, 
                           method = opt_cluster_method))
  })
})
opt_thresh_data
```

find the range of thresholds that return the correct value

```{r}
valid_thresh_range <- do.call(rbind, map(mock_meta$locus, function(l) {
  out <- range(opt_thresh_data$thresh[abs(opt_thresh_data[[l]] - nrow(mc_data)) <= 1])
  names(out) <- c('min_thresh', 'max_thresh')
  return(out)
})) %>% as_tibble()
valid_thresh_range <- as_tibble(valid_thresh_range) %>%
  mutate(locus = mock_meta$locus) %>%
  select(locus, min_thresh, max_thresh)
valid_thresh_range$min_count <- min(opt_thresh_data[mock_meta$locus])
valid_thresh_range$max_count <- max(opt_thresh_data[mock_meta$locus])
valid_thresh_range
```

Now plot

```{r}
opt_thresh_plot <- opt_thresh_data %>%
  gather(key = 'locus', value = 'otu_count', !!! mock_meta$locus) %>%
  mutate(thresh = thresh * 100) %>%
  ggplot() +
  geom_hline(yintercept = nrow(mc_data), linetype = "dashed") +
  geom_rect(aes(xmin = min_thresh * 100, xmax = max_thresh * 100, ymin = min_count, ymax = max_count), data = valid_thresh_range, alpha = 0.5) +
  geom_line(aes(x = thresh, y = otu_count)) +
  scale_x_continuous(breaks = seq(min(opt_thresh_data$thresh), max(opt_thresh_data$thresh), by = 0.01) * 100, 
                     minor_breaks = NULL) +
  scale_y_continuous(minor_breaks = 1:50) +
  facet_wrap(. ~ locus) +
  labs(x = 'PID Clustering Threshold', y = 'OTU Count', fill = '') 
ggsave(opt_thresh_plot, filename = 'optimal_clust_thresh_mock_comm.pdf', path = file.path('results'), width = 10, height = 5)
opt_thresh_plot
```

### 2021-09-23 06--mock_community.Rmd
make site public

---
title: "Mock community evaluation"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

## Prepare

### Packages used

```{r message=FALSE}
library(dplyr)
library(purrr)
library(furrr)
library(tidyr)
library(readr)
library(ggplot2)
library(sessioninfo)
library(metacoder)
library(vegan)
library(viridis)
library(taxize)
library(purrr)
library(ape)
library(ips)
library(insect)
library(phangorn)
library(DT)
library(gridExtra)
library(stringr)
library(Biostrings)
```

## Mock community reconstruction

The primary goal of any barcode is to accurately reconstruct community of organisms it is used on.
We sequenced known mock communities to evaluate how well rps10 and ITS1 can infer these communities.

### Parameters

```{r}
minimum_read_count <- 10
seed <- 1
set.seed(seed)
```

The minimum read count is higher than would be usual for a metabarcoding analysis.
We chose a relatively high read count because this analysis will only use the mock community and since the mock community is less diverse than most natural communities that metabarcoding is used on, we expected erroneous sequences to be more common than usual.

### Parallel processing

Commands that have "future" in them are run on multiple cores using the `furrr` and `future` packages.

```{r}
plan(multiprocess)
```

### Read abundance matrix and sample data

Read the sample metadata table:

```{r}
metadata <- read_csv(file.path('intermediate_data', 'metadata.csv'))
datatable(metadata)
```

Read the ASV abundance matrix:

```{r}
read_abund_mat <- function(path) {
  # Read CSV
  abundance <- read_csv(path)

  # Clean up species names
  abundance$taxonomy <- abundance$taxonomy %>%
    sub(pattern = ' aff. ', replacement = '', fixed = TRUE)

  asv_data <- parse_tax_data(abundance, class_cols = 'taxonomy', class_sep = ';',
                             class_regex = '^(.+)--(.+)--(.+)$',
                             class_key = c(taxon = 'taxon_name', boot = 'info', rank = 'taxon_rank'))
  names(asv_data$data) <- c('abund', 'score')
  transmute_obs(asv_data, 'score', sequence = sequence[input_index], boot = boot, rank = rank)
}

asv_data <- read_abund_mat(file.path('intermediate_data', 'abundance_asv.csv'))
otu_data <- read_abund_mat(file.path('intermediate_data', 'abundance_otu.csv'))
```

### Subset data to just mock community

Only `mock2` will be used in the paper. `mock1` was another, older mock community.
I will also zero out low abundance ASVs, since that would be done in a normal metabarcoding analysis.
And convert to proportions in case that is needed later.

```{r}
its_mock_sample <- metadata$sample_id[metadata$dna_type == 'mock2' & metadata$primer_pair_id == 'ITS6/7']
rps_mock_sample <- metadata$sample_id[metadata$dna_type == 'mock2' & metadata$primer_pair_id == 'rps10_Final']
mock_samples <- c(rps_mock_sample, its_mock_sample)

prep_mock_samples <- function(obj) {
  # just mock samples
  obj <- select_obs(obj, data = 'abund', sequence, taxonomy, !!! mock_samples)
  obj <- filter_obs(obj, data = 'abund', rowSums(obj$data$abund[mock_samples]) > 0, drop_taxa = TRUE)
  obj <- filter_obs(obj, data = 'score', obj$data$score$sequence %in% obj$data$abund$sequence, drop_taxa = TRUE)
  # zero out low abundance
  obj$data$abund <- zero_low_counts(obj, 'abund', min_count = minimum_read_count, other_cols = TRUE)
  obj <- filter_obs(obj, 'abund', rowSums(obj$data$abund[, mock_samples]) > 0, drop_taxa = TRUE)
  obj <- filter_obs(obj, 'score', obj$data$score$sequence %in% obj$data$abund$sequence, drop_taxa = TRUE)
  # convert to proportions
  obj$data$prop <- calc_obs_props(obj, 'abund')
  obj
}

asv_data <- prep_mock_samples(asv_data)
otu_data <- prep_mock_samples(otu_data)
print(asv_data)
print(otu_data)
```

Since there is only two samples and each ASV is only found in one:

```{r}
stopifnot(all(rowSums(asv_data$data$abund[4:5] == 0) == 1))
stopifnot(all(rowSums(otu_data$data$abund[4:5] == 0) == 1))
```

I can reformat the abundance matrix like so:

```{r}
reformat_abund_mat <- function(obj) {
  obj$data$abund <- transmute(obj$data$abund, taxon_id, sequence, taxonomy,
                              asv_count = rowSums(obj$data$abund[mock_samples]),
                              locus = ifelse(obj$data$abund[[its_mock_sample]] > 0, 'ITS', 'rps10'))
  obj$data$prop <- transmute(obj$data$prop, taxon_id, 
                             asv_count = rowSums(obj$data$prop[mock_samples]),
                             locus = ifelse(obj$data$prop[[its_mock_sample]] > 0, 'ITS', 'rps10'))
  obj
}

asv_data <- reformat_abund_mat(asv_data)
otu_data <- reformat_abund_mat(otu_data)
```

and I can add which reference sequence each asv was assigned to and its PID as seperate columns for easy use afterwards.

```{r}
add_assigned_ref_seq <- function(obj) {

  # Add name of reference sequence
  obj$data$abund$reference <- str_match(obj$data$abund$taxonomy, '.+;(.+)--[0-9]+--Reference;.+$')[,2]
  obj$data$abund$reference <- gsub(obj$data$abund$reference, pattern = '_', replacement = ' ')

  # Add PID match to reference sequence as a column
  obj$data$abund$ref_pid <- str_match(obj$data$abund$taxonomy, '.+;.+--[0-9]+--Reference;.+--([0-9.]+)--ASV$')[,2]

  obj
}

asv_data <- add_assigned_ref_seq(asv_data)
otu_data <- add_assigned_ref_seq(otu_data)
```

### Evaluating mock community reconstruction: taxonomic assignments

I will consider how the well the mock community was characterized by each method in the context of the taxonomic assignments of the ASVs by dada2.
The following metrics will be calculated:

* The number of expected mock community identified 
* The proportion of ASVs assigned to an expected member of the mock community
* The proportion of reads assigned to an expected member of the mock community

Note that all the mock community members have sequences in the reference databases:

```{r}
mc_data <- read_csv(file.path('intermediate_data', 'mock_community.csv'))
mc_syn_data <- read_csv(file.path('raw_data', 'mock_comm_synonyms.csv'))
mc_data$in_both_db <- mc_data$in_rps10_db & mc_data$in_its_db
stopifnot(all(mc_data$in_both_db))
```

To make the data easier to use, I will add the list of synonyms and the name used in the mock community to a column in the mock community data.
Note that this column is a list of character vectors.

```{r}
mc_data$all_names <- lapply(mc_data$species, function(sp_name) {
  sp_name <- gsub(sp_name, pattern = '_', replacement = ' ', fixed = TRUE)
  unique(c(sp_name, mc_syn_data$syn_name[tolower(mc_syn_data$mc_name) == tolower(sp_name)]))
})
```

I will also figure out which reference sequences might represent members of the mock community and the names of species in the mock community they correspond to.

```{r}
make_ref_data <- function(ref_path, ranks) {

  # make table from fasta info
  ref_seqs <- read_fasta(file.path('intermediate_data', 'reference_databases', ref_path))
  ref_data <- tibble(header = sub(names(ref_seqs), pattern = ';$', replacement = ''), 
                     sequence = ref_seqs)
  ref_data$header <- gsub(ref_data$header, pattern = '_', replacement = ' ')
  ref_data <- separate(ref_data, header, into = ranks, sep = ';')

  # Identify sequences that should be in the mock community
  ref_data <- mutate(
    ref_data,
    expected_in_mc = tolower(species) %in% tolower(unlist(mc_data$all_names)),
  )
  ref_data$names_in_mc <- map(ref_data$species, function(ref_sp) {
    matches_mc <- map_lgl(mc_data$all_names, function(mc_sp_names) {
      ref_sp %in% mc_sp_names
    })
    mc_data$species[matches_mc]
  })

  # Clean up species names
  ref_data$species <- ref_data$species %>%
    sub(pattern = ' subsp\\..*', replacement = '') %>%
    sub(pattern = ' var\\..*', replacement = '') %>%
    sub(pattern = ' cf\\.', replacement = '') %>%
    sub(pattern = ' aff\\.', replacement = '')

  ref_data
}

its_ref_data <- make_ref_data('its1_reference_db.fa',
                              c("domaine", "kingdom", "phylum", "class", "order", "family", "genus", "species", "reference"))
rps10_ref_data <- make_ref_data('rps10_reference_db.fa',
                                c("domaine", "kingdom", "phylum", "class", "order", "family", "genus", "species", "reference"))
```

Lets see which species in the reference database that were considered to be members of the mock community based on the synonym to make sure the synonyms are reasonable:

```{r}
print_ambiguous_mc_matches <- function(data) {
  to_look_at <- map_lgl(1:nrow(data), function(i) data$expected_in_mc[i] && (length(data$names_in_mc[[i]]) > 1 | data$species[i] != data$names_in_mc[[i]][1]))
  info <- map(which(to_look_at), function(i) paste0('Ref species name: ', data$species[i], '\nMatching mock comm names: ', paste(data$names_in_mc[[i]], collapse = ', '), '\n\n'))
  for (x in unique(info)) {
    cat(x)
  }
}
print_ambiguous_mc_matches(its_ref_data)
print_ambiguous_mc_matches(rps10_ref_data)
```

Lets get the number of OTUs and ASVs in the mock community for each locus.
I will put the results in a table of info for each locus that I will populate with additional columns throughout the analysis.

```{r}
locus_result_table <- tibble(locus = c('rps10', 'ITS'),
                             n_asv = table(asv_data$data$abund$locus)[locus],
                             n_otu = table(otu_data$data$abund$locus)[locus])
locus_result_table
```

#### The number of expected mock community species identified

I will first find which mock community members where found in the taxonomic classifications of the ASVs

```{r}
find_mock_in_tax <- function(obj, my_locus, ref_data) {
  # Get data for the locus and combine with reference data 
  names(ref_data)[colnames(ref_data) == 'sequence'] <- 'ref_sequence'
  locus_data <- filter(obj$data$abund, locus == my_locus) %>%
    left_join(ref_data, by = "reference")

  # Check which mock community members were found
  mc_tax_found <- map_lgl(mc_data$all_names, function(mc_synonyms) {
    any(mc_synonyms %in% locus_data$species)
  })

  names(mc_tax_found) <- mc_data$species
  mc_tax_found
}
```

For **rps10**:

```{r}
find_mock_in_tax(asv_data, 'rps10', rps10_ref_data)
```

For **ITS1**:

```{r}
find_mock_in_tax(asv_data, 'ITS', its_ref_data)
```

I will add that the number of mock community species found to the  table:

```{r}
locus_result_table$n_mc_in_tax <- c(sum(find_mock_in_tax(asv_data, 'rps10', rps10_ref_data)),
                                    sum(find_mock_in_tax(asv_data, 'ITS', its_ref_data)))
```

#### The proportion of the ASVs and reads assigned to members of the mock community

It would also be useful to know what proportion of the ASVs were assigned to members of the mock community:

```{r}
tax_in_mock <- function(obj, my_locus, ref_data) {
  # Get data for the locus and combine with reference data 
  names(ref_data)[colnames(ref_data) == 'sequence'] <- 'ref_sequence'
  locus_data <- filter(obj$data$abund, locus == my_locus) %>%
    left_join(ref_data, by = "reference")

  # Check which mock community members were found
  locus_data$is_mock <- locus_data$species %in% unique(unlist(mc_data$all_names))

  # Format result
  locus_data %>%
    select(species, reference, asv_count, is_mock) %>%
    arrange(desc(asv_count))
}
```

For **rps10**:

```{r}
rps10_tax_in_mock <- tax_in_mock(asv_data, 'rps10', rps10_ref_data)
as.data.frame(rps10_tax_in_mock)
```

For **ITS1**:

```{r}
its_tax_in_mock <- tax_in_mock(asv_data, 'ITS', its_ref_data)
as.data.frame(its_tax_in_mock)
```

I will add that the proportion of species found that are in the mock community to the  table:

```{r}
locus_result_table$prop_asv_in_mc_tax <- c(sum(rps10_tax_in_mock$is_mock) / nrow(rps10_tax_in_mock), 
                                           sum(its_tax_in_mock$is_mock) / nrow(its_tax_in_mock))
```

and the proportion of reads represented by those sequences

```{r}
prop_reads_in_mc <- function(aligned_data) {
  sum(aligned_data$asv_count[aligned_data$is_mock]) / sum(aligned_data$asv_count)
}

locus_result_table$prop_reads_in_mc_tax <- c(prop_reads_in_mc(rps10_tax_in_mock), 
                                             prop_reads_in_mc(its_tax_in_mock))
```

### Evaluating mock community reconstruction: sequences

Similar to the statistics generated above, I will summarize how well the mock community was reconstructed in the context of sequences found, ignoring taxonomic classifications.
Specifically, I will calculate:

* The number of expected mock community sequences present in the ASVs  
* The number of expected mock community sequences present in the ASVs, allowing for 1bp mismatch
* The proportion of ASVs matching the sequences of an expected member of the mock community
* The proportion of reads matching the sequences of an expected member of the mock community

#### The number of expected mock community sequences present in the ASVs

First, I will also count how many of the mock community species have sequences represented by at least one ASV

```{r}
# Check if bases match, allowing for IUPAC codes in reference
iupac_match <- function(asv_chars, ref_chars) {
  map2_lgl(asv_chars, ref_chars, function(asv, ref) grepl(asv, pattern = paste0('[', IUPAC_CODE_MAP[ref], ']+')))
}

# Count number of mismatches in an alignment, allowing for IUPAC codes in reference
align_mismatch <- function(alignment) {
  asv_chars <- strsplit(as.character(alignment@pattern), '')[[1]]
  ref_chars <- strsplit(as.character(alignment@subject), '')[[1]]
  sum(! iupac_match(asv_chars, ref_chars))
}

align_mock_comm_seqs <- function(obj, my_locus, ref_data) {
  # Get data for the locus and combine with reference data 
  names(ref_data)[colnames(ref_data) == 'sequence'] <- 'ref_sequence'
  locus_data <- filter(obj$data$abund, locus == my_locus) %>%
    left_join(ref_data, by = "reference")

  # Get sequences that should be in the mock community and make formatting consistent
  if (my_locus == "rps10") {
    mock_seqs <- read_fasta(file.path('raw_data', 'reference_databases', 'mock_comm_rps10_sanger.fasta'))
    names(mock_seqs) <- trimws(names(mock_seqs))
    names(mock_seqs) <- gsub(names(mock_seqs), pattern = '_', replacement = ' ')
    names(mock_seqs) <- str_match(names(mock_seqs), pattern = '^.+;(.+)$')[, 2]
    # The Pythium undulatum sequence is not complete, so I will replace it with one from a reference database
    undulatum_ref_seqs <- ref_data$ref_sequence[ref_data$species == 'Pythium undulatum']
    mock_seqs['Pythium undulatum'] <- undulatum_ref_seqs[which.max(nchar(undulatum_ref_seqs))]
  } else if (my_locus == "ITS") {
    mock_seqs <- read_fasta(file.path('raw_data', 'reference_databases', 'mock_comm_its1_sanger.fasta'))
    names(mock_seqs) <- str_match(names(mock_seqs), pattern = '^(.+) .+$')[, 2]
    names(mock_seqs) <- str_match(names(mock_seqs), pattern = '^([a-zA-Z]+ [a-zA-Z]+).*$')[, 2]
    # Plasmopara halstedii was not sequenced successfully, so I will use a reference sequence
    mock_seqs <- c(mock_seqs, setNames(its_ref_data$sequence[its_ref_data$species == "Plasmopara halstedii"], "Plasmopara halstedii"))
  } else {
    stop('Wrong locus type')
  }
  mock_seqs <- trimws(mock_seqs)

  # Align each mock community sequence and applicable reference sequences to all ASVs and return best hit
  aligned_data <-  lapply(seq_along(mock_seqs), function(i) {
    aligned <- lapply(locus_data$sequence, function(asv) pairwiseAlignment(pattern = asv, subject = mock_seqs[i], type = 'global-local'))
    tibble(
      species = names(mock_seqs)[i],
      align_len = map_dbl(aligned, nchar),
      mismatch = map_dbl(aligned, align_mismatch),
      pid = (align_len - mismatch) / align_len,
      abund = locus_data$asv_count,
      asv_seq = locus_data$sequence, 
      ref_seq = mock_seqs[i],
      alignment = aligned
    )
  })
  aligned_data <- bind_rows(aligned_data)

  aligned_data
}

filter_to_top_mc_ref_hits <- function(aligned_data) {
  aligned_data %>%
    group_by(species) %>%
    slice_max(pid, n = 1) %>%
    slice_max(abund, n = 1) %>%
    arrange(desc(abund)) %>%
    select(species, pid, mismatch, abund)
}
```

For **rps10**:

```{r}
rps10_align_data <- align_mock_comm_seqs(asv_data, 'rps10', rps10_ref_data)
as.data.frame(filter_to_top_mc_ref_hits(rps10_align_data))
```

For **ITS1**:

```{r}
its_align_data <- align_mock_comm_seqs(asv_data, 'ITS', its_ref_data)
as.data.frame(filter_to_top_mc_ref_hits(its_align_data))
```

I will also save the alignments for all of the best matches:

```{r}
print_align <- function(align) {
  ref <- as.character(align@subject)
  asv <- as.character(align@pattern)
  match_str <- paste0(ifelse(iupac_match(strsplit(asv, split = '')[[1]], strsplit(ref, split = '')[[1]]), '|', ' '), collapse = '')
  paste(
    sep = '\n',
    paste('REF:', ref), 
    paste('    ', match_str), 
    paste('ASV:', asv)
  )
}

save_best_mc_ref_alignments <- function(aligned_data, path) {
  aligned_data <- aligned_data %>%
      group_by(species) %>%
      slice_max(pid, n = 1) %>%
      slice_max(abund, n = 1)
  align_str <- map_chr(aligned_data$alignment, print_align)
  title <- paste0(aligned_data$species, ' (PID:', aligned_data$pid, ' mismatch: ', aligned_data$mismatch, '  reads: ', aligned_data$abund, ')')
  write_lines(paste0(title, '\n', align_str, '\n', sep = '\n', collapse = '\n'), file = path)
}

save_best_mc_ref_alignments(rps10_align_data, file.path('results', 'best_mc_ref_alignments_rps10.txt'))
save_best_mc_ref_alignments(its_align_data, file.path('results', 'best_mc_ref_alignments_its1.txt'))
```

and add the number of expected mock community sequences we found for each locus:

```{r}
locus_result_table$n_mc_in_seq <- c(sum(filter_to_top_mc_ref_hits(rps10_align_data)$mismatch == 0), 
                                    sum(filter_to_top_mc_ref_hits(its_align_data)$mismatch == 0))
locus_result_table$n_mc_in_seq_approx <- c(sum(filter_to_top_mc_ref_hits(rps10_align_data)$mismatch <= 1), 
                                           sum(filter_to_top_mc_ref_hits(its_align_data)$mismatch <= 1))
```

#### The proportion of ASVs matching the sequences of an expected member of the mock community

I will summarize the alignment data in the context of each ASV that best matched a reference sequence in the same way as above for the reference sequences
First, I will print out the alignment statistics:

```{r}
filter_to_top_asv_hits <- function(aligned_data) {
  aligned_data %>%
    group_by(asv_seq) %>%
    slice_max(pid, n = 1, with_ties = FALSE) %>%
    arrange(desc(abund)) %>%
    ungroup() %>%
    select(species, pid, mismatch, abund)
}
```

For **rps10**:

```{r}
as.data.frame(filter_to_top_asv_hits(rps10_align_data))
```

For **ITS1**:

```{r}
as.data.frame(filter_to_top_asv_hits(its_align_data))
```

I will also save the best alignments for each ASV:

```{r}
save_best_asv_alignments <- function(aligned_data, path) {
  aligned_data <- aligned_data %>%
      group_by(asv_seq) %>%
      slice_max(pid, n = 1, with_ties = FALSE) 
  align_str <- map_chr(aligned_data$alignment, print_align)
  title <- paste0(aligned_data$species, ' (PID:', aligned_data$pid, ' mismatch: ', aligned_data$mismatch, '  reads: ', aligned_data$abund, ')')
  write_lines(paste0(title, '\n', align_str, '\n', sep = '\n', collapse = '\n'), path = path)
}

save_best_asv_alignments(rps10_align_data, file.path('results', 'best_asv_alignments_rps10.txt'))
save_best_asv_alignments(its_align_data, file.path('results', 'best_asv_alignments_its1.txt'))
```

I will add the proportion of ASVs that matched an expected reference sequence, allowing for a 1bp difference in the alignmnet:

```{r}
prop_seq_in_mc <- function(aligned_data) {
  aligned_data <- filter_to_top_asv_hits(aligned_data)
  sum(aligned_data$mismatch <= 1) / nrow(aligned_data)
}

locus_result_table$prop_asv_in_mc_seq <- c(prop_seq_in_mc(rps10_align_data), prop_seq_in_mc(its_align_data))
```

and the proportion of reads represented by those sequences

```{r}
prop_reads_in_mc <- function(aligned_data) {
  aligned_data <- filter_to_top_asv_hits(aligned_data)
  sum(aligned_data$abund[aligned_data$mismatch <= 1]) / sum(aligned_data$abund)
}

locus_result_table$prop_reads_in_mc_seq <- c(prop_reads_in_mc(rps10_align_data), prop_reads_in_mc(its_align_data))
```

Finally, I will save the table of summary statistics generated in the last few sections:

```{r}
num_col <- map_lgl(locus_result_table, is.numeric)
locus_result_table[num_col] <- lapply(locus_result_table[num_col], round, digits = 3)
write_csv(locus_result_table, file = file.path('results', 'mock_community_summary_statistics.csv'))
print(locus_result_table)
```

### Plot composition of the mock community

I will make heat trees for  each of the mock community samples with the taxa that are should not be there highlighted. 
Here is a function to calculated the data plotted for each sample:

```{r message=FALSE}
plot_comp <- function(locus, title, path) {
  x <- asv_data$clone(deep = TRUE)

  # Get right reference data for locus
  if (locus == 'rps10') {
    ref_data <- rps10_ref_data
  } else if (locus == 'ITS') {
    ref_data <- its_ref_data
  } else {
    stop('Invalid locus')
  }

  # Subset to only samples used in this plot
  x$data$prop <- x$data$prop[x$data$prop$locus == locus, ]
  x$data$abund <- x$data$abund[x$data$abund$locus == locus, ]
  x$data$score <- x$data$score[x$data$score$sequence %in% x$data$abund$sequence, ]

  # Get per-taxon mean proportions and read counts
  x$data$tax_prop <- calc_taxon_abund(x, 'prop', cols = 'asv_count', out_names = 'tax_prop')
  x$data$tax_abund <- calc_taxon_abund(x, 'abund', cols = 'asv_count', out_names = 'tax_count')

  # Remove taxa and bootstrap scores for taxa/ASVs not in plot
  x <- filter_taxa(x, tax_count > minimum_read_count, reassign_obs = FALSE)

  # Get mean bootstrap score per taxon
  x$data$score$boot[x$data$score$rank == 'ASV'] <- NA
  x$data$tax_prop$mean_boot <- obs_apply(x, 'score', value = 'boot', recursive = FALSE, func = function(boots) {
    mean(as.numeric(boots), na.rm = TRUE)
  }) %>% unlist

  # Replace the bootstrap value for the ASVs with their PID to the reference
  x$data$abund$assigned_pid <- as.numeric(str_match(x$data$abund$taxonomy, pattern = '^.+--([0-9.]+)--ASV$')[,2]) / 100
  x$data$tax_prop$mean_boot[match(x$data$abund$taxon_id, x$data$tax_prop$taxon_id)] <- x$data$abund$assigned_pid * 100

  # Find which taxa are part of mock community
  species_ids <- map_chr(supertaxa(x, value = 'taxon_ids', recursive = 2), `[`, 2)[x$data$abund$taxon_id]
  species_found <- gsub(taxon_names(x)[species_ids], pattern = '_', replacement = ' ')
  species_in_mc <- tolower(species_found) %in% tolower(unlist(mc_data$all_names))
  x$data$taxa_in_mc <- map_lgl(subtaxa(x, value = 'taxon_ids', include_input = TRUE), function(i) any(i %in% species_ids[species_in_mc]))

  # Plot
  x %>% 
    filter_taxa(! is_stem) %>%
    # filter_taxa(taxon_ranks != 'Reference') %>%
    # remove_redundant_names() %>%
    heat_tree(node_label = ifelse(taxon_ranks == 'ASV', 
                                  paste0('PID: ', format(mean_boot, digits = 3, trim = TRUE), '\nN: ', tax_count),
                                  gsub(taxon_names, pattern = '_', replacement = ' ')),
              node_label_color = ifelse(taxa_in_mc | taxon_ranks %in% c('ASV', 'Reference'), 'black', 'darkred'),
              node_size = tax_prop, 
              node_size_range = c(0.005, 0.035),
              node_size_interval = c(0, 1),
              node_label_size_range = c(0.012, 0.025),
              node_size_axis_label = 'Mean read proportion',
              node_color = ifelse(taxon_ranks == 'ASV', 'grey', mean_boot),
              node_color_axis_label = 'Mean bootstrap',
              node_color_trans = 'linear',
              node_color_range = c('red', 'red', 'red', 'red', 'orange', 'yellow', 'yellow', 'green', 'lightblue'),
              node_color_interval = c(0, 100),
              title = title,
              output_file = path)
}

sample_plots <- lapply(c("rps10", "ITS"), function(locus) {
  mdat <- metadata[metadata$locus == locus & metadata$sample_id %in% mock_samples, ]
  mdat$primer_pair_id <-  gsub(mdat$primer_pair_id, pattern = '/', replacement = '-', fixed = TRUE)
  my_plot <- plot_comp(locus = locus, 
                       title = locus,
                       path = file.path('results', paste0('mock_comm_heat_tree_', mdat$primer_pair_id, '.pdf')))
  print(my_plot)
  my_plot
})
```

## Phylogenetic tree

```{r}
make_full_tree <- function(locus, ref_data, path, title = locus, only_best_matches = FALSE, legend = TRUE) {

  # Subset data to just locus plotted
  asv_match_data <- filter(asv_data$data$abund, locus == !! locus)

  # Only consider the ASVs in each species with the best match to the reference
  if (only_best_matches) {
    asv_match_data <- map(split(asv_match_data, asv_match_data$species), 
                          function(x) arrange(x, desc(assigned_pid), desc(asv_count))[1, ]) %>%
      bind_rows()
  }

  # Get sequences of mock community members
  if (locus == "rps10") {
    mock_seqs <- read_fasta(file.path('raw_data', 'reference_databases', 'mock_comm_rps10_sanger.fasta'))
    names(mock_seqs) <- trimws(names(mock_seqs))
    names(mock_seqs) <- gsub(names(mock_seqs), pattern = '_', replacement = ' ')
    names(mock_seqs) <- str_match(names(mock_seqs), pattern = '^.+;(.+)$')[, 2]
  } else if (locus == "ITS") {
    mock_seqs <- read_fasta(file.path('raw_data', 'reference_databases', 'mock_comm_its1_sanger.fasta'))
    names(mock_seqs) <- str_match(names(mock_seqs), pattern = '^(.+) .+$')[, 2]
  } else {
    stop('Wrong locus type')
  }

  # Combine sequences
  seq_data <- tibble(
    name = c(paste0('ASV (', asv_match_data$asv_count, ')'), names(mock_seqs)),
    type = rep(c('ASV', 'Reference'), c(nrow(asv_match_data), length(mock_seqs))),
    sequence = c(asv_match_data$sequence, mock_seqs)
    )
  seq_data$sequence <- trimws(toupper(seq_data$sequence))

  # Align sequences:
  aligned <- seq_data$sequence %>%
    setNames(1:nrow(seq_data)) %>%
    char2dna() %>%
    mafft(method = 'localpair', exec = 'mafft')

  # Make tree
  make_tree <- function(aligned) {
    dist <- dist.dna(aligned, model = 'N')
    tree <- nj(dist)
    tree <- ladderize(tree)
    tree <- midpoint(tree)
    tree
  }
  tree <- make_tree(aligned)
  boot <- boot.phylo(tree, aligned, make_tree, quiet = TRUE)
  # boot[boot == 100] <- NA

  # Plot tree
  # pdf(file = path, width = 7, height = 8)
  tree$tip.label <- seq_data$name
  tree$node.label <- boot
  plot.phylo(tree)
  title(main = title, adj = 0)
  nodelabels(ifelse(boot == 100, NA, boot), frame = 'none', cex = 0.61, adj = c(1.5, -.2))
  axis(side = 1)
  title(xlab = 'Number of differing sites')
  # dev.off()

  return(boot)
}

make_phylo_plot <- function() {
  par(mfrow = c(1, 2))
  its1_phylo_boot <- make_full_tree(locus = 'ITS',
                                    ref_data = its_ref_data,
                                    path = file.path('results', 'mock_comm_phylo_full_its1.pdf'),
                                    title = 'A. ITS1',
                                    legend = TRUE)
  rps10_phylo_boot <- make_full_tree(locus = 'rps10', 
                                     ref_data = rps10_ref_data, 
                                     path = file.path('results', 'mock_comm_phylo_full_rps10.pdf'),
                                     title = 'B. RPS10',
                                     legend = FALSE)
  tibble(locus = rep(c('rps10', 'ITS'), c(length(rps10_phylo_boot), length(its1_phylo_boot))),
         boot = c(rps10_phylo_boot, its1_phylo_boot),
         rank = 'Branch',
         type = 'Neighbor Joining')

}
```

Plot and save the plot:

```{r fig.width=16, fig.height=16}
pdf(file = file.path('results', 'mock_comm_phylo_full.pdf'), width = 16, height = 16)
phylo_boot_data <- make_phylo_plot()
dev.off()
phylo_boot_data <- make_phylo_plot() 
```

### Bootstrap analysis

In order to see how well the two barcodes can be used to assign taxonomy and resolve phylogenetic trees, I will compare the bootstrap values from dada2 and the neighbor-joining trees produced with ape.

```{r}
tax_boot_data <- asv_data$data$score[asv_data$data$score$rank %in% c('Genus', 'Species', 'Reference'), ]
# tax_boot_data <- asv_data$data$score
tax_boot_data$locus <- asv_data$data$abund$locus[match(tax_boot_data$sequence, asv_data$data$abund$sequence)]
tax_boot_data$type <- 'Naive Bayesian Classifier'
tax_boot_data$boot <- as.integer(tax_boot_data$boot)
boot_data <- bind_rows(phylo_boot_data, tax_boot_data[, colnames(phylo_boot_data)])
boot_data
```

Now I can graph it with histograms

```{r fig.width=5, fig.height=5}
boot_plot <- boot_data %>%
  # filter(type == 'Taxonomic') %>%
  mutate(rank = factor(rank, levels = rev(c('Branch', 'Genus', 'Species', 'Reference')), ordered = TRUE),
         locus = ordered(c(rps10 = 'rps10', ITS = 'ITS1')[locus], levels = c('rps10', 'ITS1'))) %>%
  ggplot(aes(x = rank, y = boot, fill = locus)) +
  facet_grid(. ~ type, scales = 'free', space = 'free_x') +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.8, end = 0.2) +
  labs(x = '', y = 'Bootstrap Score', fill = 'Locus') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position="bottom") 
boot_plot
ggsave(boot_plot, filename = 'mock_comm_bootstrap.pdf', path = 'results', width = 7, height = 5)
```

**Caption:**

Figure #: The distribution of bootstrap scores for the taxonomic assignment of ASVs in the mock community for the ITS1 and Rps10 loci. The RDP Naive Bayesian Classifier "Reference", "Species", and "Genus" scores refer the ability to consistently assign ASVs to a particular reference sequence, species, or genus respectively when the data is resampled. The neighbor joining tree scores quantify how consistent the branching pattern of the resulting tree is when the data is resampled.

## Software used

```{r}
sessioninfo::session_info()
```

### 2021-09-23 07--nontarget_amplification.Rmd
make site public

---
title: "Non-target amplification"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

## Prepare

### Packages used

```{r message=FALSE}
library(dplyr)
library(purrr)
library(furrr)
library(tidyr)
library(readr)
library(ggplot2)
library(sessioninfo)
library(metacoder)
library(vegan)
library(viridis)
library(DT)
library(stringr)
library(qsubmitter)
library(taxize)
library(forcats)
library(ips)
library(insect)
library(ape)
library(phangorn)
```

### Parameters

```{r}
minimum_read_count <- 10
seed <- 1
set.seed(seed)
```

## Detecting non-target amplification

Since the reference databases will not have all the non-targets, it will be hard to detect non-target amplification.
I will use blast against Genbank nt.
Genbank is not curated, but it should be good enough to reliably assign a phylum.

```{r}
abundance_asv <- read_csv(file.path('intermediate_data', 'abundance_asv.csv'))
abundance_asv
abundance_otu <- read_csv(file.path('intermediate_data', 'abundance_otu.csv'))
abundance_otu
metadata <- read_csv(file.path('intermediate_data', 'metadata.csv'))
metadata
```

### Select only environmental samples

Non-target DNA is likely to be found the in the environmental samples, so I will restrict this analysis to those.

```{r}
metadata <- filter(metadata,
                   dna_type %in% c("leaf", "WR.soil", "ag.soil", "drip", "Pan.soil"),
                   primer_pair_id %in% c('rps10_Final', 'ITS6/7'))
metadata
abundance_asv <- select(abundance_asv, sequence:taxonomy, !!! metadata$sample_id)
abundance_asv
abundance_otu <- select(abundance_otu, sequence:taxonomy, !!! metadata$sample_id)
abundance_otu
```

### Filtering out low abundance_asv ASVs

Most low-abundance_asv ASVs would normally be filtered out during most analyses, so I will do that to.

```{r}
abundance_asv <- abundance_asv[rowSums(abundance_asv[, metadata$sample_id]) >= minimum_read_count, ]
abundance_asv
abundance_otu <- abundance_otu[rowSums(abundance_otu[, metadata$sample_id]) >= minimum_read_count, ]
abundance_otu
```

### Group ASVs into target and non-target

```{r}
group_key <- c(Oomycetes = 'Oomycetes|Oomycota', Fungi = 'Fungi', Bacteria = 'Bacteria', Plant = 'Viridiplantae', 
               Archaea = 'Archaea', Animals = 'Metazoa', Virus = 'Viruses', Protist = 'Jakobida|Alveolata|Euglenozoa|Amoebozoa')

groups_target_nontarget <- function(abundance) {
  map_chr(abundance$blast_tax, function(x) {
    if (is.na(x)) {
      return('Unknown')
    }
    is_org <- map_lgl(group_key, grepl, x = x)
    if (sum(is_org) > 1) {
      stop('multiple matches')
    } else if (sum(is_org) == 1) {
      return(names(group_key)[is_org])
    } else {
      return('Other')
    }
  })
}
abundance_asv$group <- groups_target_nontarget(abundance_asv)
abundance_otu$group <- groups_target_nontarget(abundance_otu)
```

### Tally number of reads/ASVs per sample in each group

I will make a table with the numbers of ASVs, reads, and the proportion of reads for each group.
First the ASV counts:

```{r}
groups <- unique(abundance_asv$group)

count_seqs <- function(abundance, measure) {
  counts <- map(metadata$sample_id, function(id) {
    map_dbl(groups, function(group) {
      is_group <- abundance$group == group & abundance[[id]] > 0
      sum(is_group)
    })
  })
  names(counts) <- metadata$sample_id
  counts <- as_tibble(c(group = list(groups), 
                        measure = list(rep(measure, length(group_key))),
                        counts))
  counts
}
asv_counts <- count_seqs(abundance_asv, 'ASVs')
otu_counts <- count_seqs(abundance_otu, 'OTUs')
```

Then read counts

```{r}
count_reads <- function(abundance) {
  read_counts <- map(metadata$sample_id, function(id) {
    map_dbl(groups, function(group) {
      is_group <- abundance$group == group & abundance[[id]] > 0
      sum(abundance[[id]][is_group])
    })
  })
  names(read_counts) <- metadata$sample_id
  read_counts <- as_tibble(c(group = list(groups), 
                             measure = list(rep('read count', length(group_key))),
                             read_counts))
  read_counts
}
read_counts_asv <- count_reads(abundance_asv)
read_counts_otu <- count_reads(abundance_otu)
```

And then I can convert those count to proportions:

```{r}
convert_to_prop <- function(read_counts, abundance) {
  read_props <- read_counts
  read_props[metadata$sample_id] <- map(metadata$sample_id, function(id) {
    read_props[[id]] / sum(abundance[[id]])
  })
  read_props$measure <- 'Reads'
  read_props
}
read_props_asv <- convert_to_prop(read_counts_asv, abundance_asv)
read_props_otu <- convert_to_prop(read_counts_otu, abundance_otu)
```

### Plot barchart of counts

Finally I can combine these into a single table and add some metadata

```{r}
group_abund <- bind_rows(otu_counts, asv_counts, read_counts_asv, read_props_asv) %>%
  gather(key = 'sample_id', value = 'abund', !!! metadata$sample_id) %>%
  left_join(metadata, by = 'sample_id')
group_abund
```

I will calculate the proportion of ASVs and OTUs assigned to each group for each locus

```{r}
group_props <- group_abund %>%
  group_by(measure, locus) %>%
  mutate(sum_abund = sum(abund)) %>%
  group_by(group, .add = TRUE) %>%
  summarise(prop = round(sum(abund) / sum_abund[1], digits = 4)) %>%
  filter(measure %in% c("ASVs", "OTUs", "read count")) 
datatable(group_props)
```

Check that proportions sum to 1 (there will be a bit of rounding error):

```{r}
group_props %>%
  summarise(sum = sum(prop))
```

Lets take a look at what proportion of oomycetes was found for each metric:

```{r}
filter(group_props, group == "Oomycetes")
```

Finally, lets plot this information:

```{r fig.width=5, fig.height=5}
nontarget_plot <- group_abund %>% 
  filter(measure %in% c('Reads', 'ASVs', 'OTUs')) %>%
  mutate(dna_type = fct_collapse(dna_type,
                                 Water = c('drip'),
                                 Plant = c('leaf'),
                                 Soil = c('ag.soil', 'Pan.soil', 'WR.soil')),
         measure = factor(measure, levels = c('Reads', 'ASVs', 'OTUs'), ordered = TRUE),
         locus = ordered(c(rps10 = 'rps10', ITS = 'ITS1')[locus], levels = c('rps10', 'ITS1'))) %>%
  group_by(measure, locus, dna_type) %>%
  mutate(abund = abund / sum(abund)) %>%
  mutate(group = fct_collapse(group,
                              Oomycetes = c('Oomycetes'),
                              Fungi = c('Fungi'),
                              Unknown = c('Unknown'),
                              Other = c('Protist', 'Other', 'Plant', 'Bacteria', 'Animals'))) %>%
  mutate(group = factor(group, levels = c("Unknown", "Other", "Fungi", "Oomycetes"), ordered = TRUE)) %>%
  # ggplot(aes(x = locus, y = abund, fill = group)) +
  ggplot(aes(x = dna_type, y = abund, fill = group)) +
  geom_bar(stat = 'identity') +
  # facet_grid(measure ~ dna_type) +
  facet_grid(measure ~ locus) +
  scale_fill_viridis_d(begin = 0.8, end = 0.2) +
  labs(x = NULL, y = 'Proportion', fill = NULL) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = 'bottom') 
ggsave(nontarget_plot, filename = 'non_target_amplification.pdf', path = 'results', width = 5, height = 5)
nontarget_plot
```

Caption:

Figure #: Target vs non-target amplification using oomycete-specific primers for the ITS1 and RPS10 loci. Reads and ASVs are from a variety of environmental samples, grouped into soil, water, and plant tissue samples. ASV sequences were given a coarse taxonomic assignment based on BLAST searches against the NCBI nucleotide sequence database. Those assigned "Unknown" did not have a match with a E-value of at least 0.001. Sequences in "Other" include plant, animal, bacterial, and protist sequences.

## Look at unknown rps10 sequences

```{r}
unknown_asv_data <- abundance_asv %>% 
  gather(key = 'sample_id', value = 'count', !!! metadata$sample_id) %>%
  left_join(metadata, by = 'sample_id') %>%
  filter(locus == 'rps10', group == 'Unknown', count > 0) %>%
  select(sequence, sample_id, count, dna_type, sample_type)
datatable(unknown_asv_data)
```

It seems a lot these are short:

```{r}
hist(nchar(unknown_asv_data$sequence))
```

And somewhat low abundance:

```{r}
hist(unknown_asv_data$count)
```

The really short and low abundance ones are probably errors, so lets remove those:

```{r}
unknown_asv_data <- filter(unknown_asv_data, nchar(sequence) >= 100, count >= 100)
```

I will get a random sample of oomycete reference sequences and see if they cluster together with any in particular or form separate clades

```{r}
rps10_seqs <- read_fasta(file.path('intermediate_data', 'reference_databases', 'rps10_reference_db.fa'))
names(rps10_seqs) <- gsub(names(rps10_seqs), pattern = ';oodb_[0-9]+;$', replacement = '') %>%
  gsub(pattern = '^.+;', replacement = '') %>%
  gsub(pattern = '_', replacement = ' ')
ref_subsample <- rps10_seqs[!duplicated(names(rps10_seqs))]
ref_subsample <- ref_subsample[sample(length(ref_subsample), 10)]
```

And make a multiple sequence alignment with both the reference sequences and unknown sequences

```{r}
unknown_asv_seqs <- setNames(unknown_asv_data$sequence, paste0('ASV ', seq_along(unknown_asv_data$sequence), ' (', unknown_asv_data$count, ')'))
aligned <- c(ref_subsample, unknown_asv_seqs) %>%
  char2dna() %>%
  mafft(method = 'localpair', exec = 'mafft')
```

It turns out that the unknown sequences are very different from eachother and the reference sequences; so much so that I could not calculate a distance matrix to make a tree.

```{r}
image(aligned)
```

I manually BLASTed a few of the sequences on NCBI's website to verify the BLAST on all of the ASVs worked properly and did not find any close matches.
These seem to be some kind of errors or mispriming against unknown organisms.

## Software used

```{r}
sessioninfo::session_info()
```

### 2021-09-23 08--amplicon_taxonomic_resolution.Rmd
make site public

---
title: "Taxonomic resolution"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

## Prepare

### Packages used

```{r message=FALSE}
library(dplyr)
library(purrr)
library(furrr)
library(tidyr)
library(readr)
library(ggplot2)
library(sessioninfo)
library(metacoder)
library(vegan)
library(viridis)
library(DT)
library(stringr)
library(qsubmitter)
library(taxize)
library(forcats)
library(Biostrings)
library(ips)
library(insect)
library(ggpubr)
library(knitr)
```

### Parameters

```{r}
seed <- 1
only_shared_species <- TRUE # If TRUE, only use species that are present in both databases
subsample_to_same_num <- TRUE # If TRUE, subsample the amplicons used to the same number for each database
set.seed(seed)
```

### Parallel processing

Commands that have "future" in them are run on multiple cores using the `furrr` and `future` packages.

```{r}
plan(multiprocess)
```

### Introduction

This analysis will attempt to estimate how good each locus is at distinguishing species, what I am calling "taxonomic resolution".
I will use the primer sequences and reference databases for each method to extract the portion of the sequence that would be amplified.
I then align each sequence to all other sequences and record how many base pairs different each sequence is from the most similar sequence from a different species.
The method with the highest numbers should be the one with the highest taxonomic resolution.
This is similar to a barcode gap analysis, except that only the most similar comparisons are considered.

### Extract all oomycete from reference databases

```{r}
rps_ref <- read_fasta(file.path('intermediate_data', 'reference_databases', 'rps10_reference_db.fa'))
its_ref <- read_fasta(file.path('intermediate_data', 'reference_databases', 'its1_reference_db.fa'))
rps_ref <- rps_ref[grepl(names(rps_ref), pattern = 'Oomycetes')]
its_ref <- its_ref[grepl(names(its_ref), pattern = 'Oomycetes')]
```

Remove spaces from sequences (not sure why there would be spaces).

```{r}
rps_ref <- trimws(gsub(rps_ref, pattern = ' ', replacement = ''))
its_ref <- trimws(gsub(its_ref, pattern = ' ', replacement = ''))
```

Replace names with just species names

```{r}
names(rps_ref) <- str_match(names(rps_ref), ';([a-zA-Z0-9_\\-\'". ]+);oodb_.+;$')[,2]
names(its_ref) <- str_match(names(its_ref), ';([a-zA-Z0-9_\\-\'". ]+);(unite|rob2011|phytodb|mock)_.+;$')[,2]
```

Replace underscores with spaces

```{r}
names(rps_ref) <- gsub(names(rps_ref), pattern = '_', replacement = ' ')
names(its_ref) <- gsub(names(its_ref), pattern = '_', replacement = ' ')
```

Remove species with ambiguous names or numbers

```{r}
is_ambiguous <- function(n) {
  grepl(n, pattern = ' sp$') |
    grepl(n, pattern = ' spp$') | 
    grepl(n, pattern = ' spp.') |
    grepl(n, pattern = ' sp.') |
    grepl(n, pattern = ' aff. ')  | 
    grepl(n, pattern = ' cf.') |
    grepl(n, pattern = '[0-9]+') 
}
rps_ref <- rps_ref[! is_ambiguous(names(rps_ref))]
its_ref <- its_ref[! is_ambiguous(names(its_ref))]
```

Remove information below species

```{r}
names(rps_ref) <- sub(names(rps_ref), pattern = ' var\\.?.+$', replacement = '')
names(rps_ref) <- sub(names(rps_ref), pattern = ' subsp\\.?.+$', replacement = '')
names(its_ref) <- sub(names(its_ref), pattern = ' var\\.?.+$', replacement = '')
names(its_ref) <- sub(names(its_ref), pattern = ' subsp\\.?.+$', replacement = '')
```

Now the names should be cleaned up enough that they can be compared to eachother:

```{r}
unique(names(rps_ref))
unique(names(its_ref))
```

Although there are a similar number of species, each database has a distinct set of species:

```{r}
table(unique(names(rps_ref)) %in% unique(names(its_ref)))
table(unique(names(its_ref)) %in% unique(names(rps_ref)))
```

This could influence the comparison, so we can use only species present in both to avoid that:

```{r}
if (only_shared_species) {
  common_species <- intersect(unique(names(rps_ref)), unique(names(its_ref)))
  rps_ref <- rps_ref[names(rps_ref) %in% common_species]
  its_ref <- its_ref[names(its_ref) %in% common_species]
}
```

### Extract predicted amplicons

Only the amplified sequence contribute to the taxonomic resolution of the locus, so I will have to find that region in reference database sequences.
Not all reference sequences will have the entire region however.
Some will have the region but not the primer sites, making things more complicated.

```{r}
primer_data <- read_csv(file.path("raw_data", "primer_data.csv"))
primer_data <- filter(primer_data, primer_id != 'rps10_f-F')
primer_data
```

The `matchProbePair` function from the `Biostrings` package can extract a region predicted to be amplified by primers, but for some reason it does not allow for ambiguity codes, even though it is a trivial change to make it do so.
Below is a modified version of the function that does allow ambiguity codes by providing the `fixed` parameter and passing it to the `matchPattern` calls.

```{r}
matchProbePair <- function(Fprobe, Rprobe, subject, algorithm="auto",
             logfile=NULL, verbose=FALSE, fixed = FALSE)
    {
        ## This won't copy the data if Fprobe and Rprobe are already DNAString objects
        F <- DNAString(Fprobe)
        R <- DNAString(Rprobe)

        ## F and R hits on the + strand
        Fp_hits <- start(matchPattern(F, subject, algorithm=algorithm, fixed = fixed))
        Rp_hits <- start(matchPattern(R, subject, algorithm=algorithm, fixed = fixed))

        ## F and R hits on the - strand
        Fm_hits <- end(matchPattern(reverseComplement(F), subject, algorithm=algorithm, fixed = fixed))
        Rm_hits <- end(matchPattern(reverseComplement(R), subject, algorithm=algorithm, fixed = fixed))

        if (verbose) {
            cat("Fp_hits:", Fp_hits, "  Rp_hits:", Rp_hits,
                "  Fm_hits:", Fm_hits, "  Rm_hits:", Rm_hits, "\n")
        }

        matches0 <- Biostrings:::reduceProbePairMatches(c(Fp_hits, Rp_hits), c(Fm_hits, Rm_hits))
        ans <- Views(subject, start=matches0$start, end=matches0$end)

        if (!is.null(logfile)) {
            nFp <- length(Fp_hits)
            nRp <- length(Rp_hits)
            nFm <- length(Fm_hits)
            nRm <- length(Rm_hits)
            nmatches0 <- length(ans)
            ## cat("", ..., sep="\t") is a trick to get an extra tab
            cat("", nFp, nRp, nFm, nRm, nmatches0, file=logfile, sep="\t")
        }
        ans
}
```

The code below extracts predicted amplicons using primer matches, removes the primers, and aligns the trimmed amplicons to the sequences that did not have primer matches to try to find other sequences in the reference databases that do not have primer sequences but are full length matches to those that do.

```{r}
get_amplicon_chr <- function(seq) {
  map2_chr(seq@ranges@start, seq@ranges@width, function(s, w) {
    as.character(seq@subject[seq(from = s, length.out = w)])
  })
}

calc_amplicons <- function(seqs, forward, reverse, min_coverage = 0.9, ...) {
  # Get simulated amplicons using primers
  full_amps <- future_map(seqs, function(s) {
    get_amplicon_chr(matchProbePair(DNAString(s), Fprobe = forward, Rprobe = reverse))
  })

  # Check for multiple possible amplicons per input
  if (any(map_int(full_amps, length) > 1)) {
    stop('Some inputs have more than one amplicons.')
  }
  full_amps <- unlist(full_amps[map_int(full_amps, length) == 1])

  # Remove primers from amplicons
  full_amps <- substr(full_amps, start = nchar(forward) + 1, stop = nchar(full_amps) - nchar(reverse))

  # Align unamplified sequences with best matching amplicons
  unamped <- seqs[! names(seqs) %in% names(full_amps)]
  unamped_aligned <- future_map_chr(unamped, function(s) {
    aligned <- pairwiseAlignment(pattern = full_amps, subject = s, 
                                 type = 'global-local',
                                 # type = 'overlap',
                                 gapOpening = 10, gapExtension = 4)
    best_score_i <- which.max(aligned@score)
    best_align <- aligned[best_score_i]
    aligned_subject_char <- gsub(as.character(best_align@subject), pattern = '-', replacement = '')
    aligned_amp_char <- gsub(as.character(best_align@pattern), pattern = '-', replacement = '')
    if (best_align@pattern@range@width < best_align@pattern@unaligned@ranges@width * min_coverage || nchar(aligned_subject_char) == 0) {
      aligned_subject_char <- NA
    }
    aligned_subject_char
  })

  # Add full amplicons to inferred amplicons from alignment
  all_amps <- c(full_amps, unamped_aligned[! is.na(unamped_aligned)])
  all_amps <- all_amps[is.na(all_amps) | nchar(all_amps) > 0]
  setNames(all_amps[names(seqs)], names(seqs))
}

rps_amps <- calc_amplicons(rps_ref, 
                           forward = primer_data$sequence[primer_data$direction == 'Forward' & primer_data$locus == 'rps10'],
                           reverse = primer_data$sequence[primer_data$direction == 'Reverse' & primer_data$locus == 'rps10'])
its_amps <- calc_amplicons(its_ref, 
                           forward = primer_data$sequence[primer_data$direction == 'Forward' & primer_data$locus == 'ITS'],
                           reverse = primer_data$sequence[primer_data$direction == 'Reverse' & primer_data$locus == 'ITS'])
```

The amplicon could not be extracted from many sequences, probably because it was incomplete, but enough can be to compare the two methods.
Here is the proportion of reference sequences with full length amplicons extracted:

```{r}
table(is.na(rps_amps))
table(is.na(its_amps))
```

Lets remove those from the databases:

```{r}
rps_amps <- rps_amps[! is.na(rps_amps)]
its_amps <- its_amps[! is.na(its_amps)]
```

Lets also remove any sequences that are duplicates (same sequence and same species name):

```{r}
rps_amps <- rps_amps[! duplicated(paste0(names(rps_amps), rps_amps))]
its_amps <- its_amps[! duplicated(paste0(names(its_amps), its_amps))]
```

And sort them by name:

```{r}
rps_amps <- rps_amps[order(names(rps_amps))]
its_amps <- its_amps[order(names(its_amps))]
```

And here is the distribution of amplicon lengths with and without primers:

```{r fig.width=10, fig.height=5}
amp_data <- tibble(locus = rep(c('rps10', 'ITS'), c(length(rps_amps), length(its_amps))),
                   seq = c(rps_amps, its_amps),
                   name = names(c(rps_amps, its_amps)),
                   without_primers = nchar(seq))
amp_data$locus <- ordered(amp_data$locus, levels = c('rps10', 'ITS')) # order
its_primer_length <- sum(nchar(primer_data$sequence[primer_data$locus == 'ITS']))
rps_primer_length <- sum(nchar(primer_data$sequence[primer_data$locus == 'rps10']))
amp_data <- mutate(amp_data, with_primers = without_primers + ifelse(locus == 'ITS', its_primer_length, rps_primer_length))

amp_len_plot <- amp_data %>%
  gather(key = 'type', value = 'length', without_primers, with_primers) %>%
  mutate(type = c(without_primers = 'Without primers', with_primers = 'With primers')[type]) %>%
  ggplot(aes(x = length)) +
  geom_histogram(binwidth = 1) +
  facet_grid(type ~ locus, scales = 'free_x') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()) 
amp_len_plot
ggsave(amp_len_plot, filename = 'predicted_amplicon_length.pdf', path = file.path('results'), width = 10, height = 5)
```

Lets look at which species have the largest amplicons:

```{r}
amp_data %>%
  filter(locus == 'rps10', with_primers > 480) %>%
  arrange(desc(with_primers)) %>%
  select(name, without_primers, with_primers, seq) %>%
  group_by(name) %>%
  filter(with_primers == max(with_primers)) %>%
  filter(! duplicated(name)) %>%
  datatable()
```

and which have the shortest:

```{r}
amp_data %>%
  filter(locus == 'rps10', with_primers < 460) %>%
  arrange(with_primers) %>%
  select(name, without_primers, with_primers, seq) %>%
  group_by(name) %>%
  filter(with_primers == max(with_primers)) %>%
  filter(! duplicated(name)) %>%
  datatable()
```

I will save FASTA files of the amplicons for use in other analyses

```{r}
paste0(">", names(rps_amps), "\n", rps_amps) %>%
  write_lines(file = file.path('intermediate_data', 'reference_databases', 'predicted_amplicons_rps10.fa'))
paste0(">", names(its_amps), "\n", its_amps) %>%
  write_lines(file = file.path('intermediate_data', 'reference_databases', 'predicted_amplicons_its1.fa'))
```

I will also save a CSV version of this:

```{r}
amp_data %>%
  filter(locus == 'rps10') %>%
  arrange(desc(with_primers)) %>%
  select(name, without_primers, with_primers, amplicon = seq) %>%
  group_by(name) %>%
  filter(with_primers == max(with_primers)) %>%
  filter(! duplicated(name)) %>%
  write_csv(file = file.path('results', 'rps10_predicted_amplicon_length.csv'))
```

Again, lets only use species present in both

```{r}
if (only_shared_species) {
  common_species <- intersect(unique(names(rps_amps)), unique(names(its_amps)))
  rps_amps <- rps_amps[names(rps_amps) %in% common_species]
  its_amps <- its_amps[names(its_amps) %in% common_species]
}
unique(names(rps_amps))
```

### Align predicted amplicons

To make a distance matrix, I will align the amplicons:

```{r fig.height=15}
nj_tree <- function(seqs, ...) {
  # Align sequences:
  aligned <- seqs %>%
    insect::char2dna() %>%
    ips::mafft(method = 'localpair', exec = '/usr/bin/mafft')

  # Make distance matrix
  dist <- ape::dist.dna(aligned, ...)

  # Make tree
  tree <- ape::nj(dist)
  tree <- ape::ladderize(tree)
  tree <- phangorn::midpoint(tree)

  tree
}

calc_align <- function(amps) {
  amps %>%
    gsub(pattern = '[^AGCTN-]', replacement = 'N') %>%
    char2dna() %>%
    mafft(method = "globalpair", exec = 'mafft')
}

plot_align <- function(amps, aligned, title, spot = "top", start = 1, end = ncol(aligned)) {
  # aligned <- rps_amp_aligned; amps <- rps_amps; title <- "Rps10"; spot = "top"; start = 50; end = ncol(aligned)

  # Subset alignment to region to plot       horiz = TRUE, xpd = TRUE)
  aligned <- aligned[, start:end]

  # Make tree and order alignments to match tree
  tree <- nj_tree(amps)
  tree <- ladderize(tree)
  aligned <- aligned[rev(tree$tip.label[tree$edge[tree$edge[,2] <= length(tree$tip.label), 2]]), ]

  # Get sequence conservation at each position 
  major_allele_prop <- unlist(lapply(1:ncol(aligned), function(index) {
    chars <- as.character(aligned[, index, drop = TRUE])
    most_common_base <- sort(table(chars), decreasing = TRUE)[1]
    return(most_common_base / nrow(aligned))
  }))
  barplot_data <- data.frame(prop = major_allele_prop,
                             fill = 1 - major_allele_prop)

  # Get consensus sequence
  consensus <- as.DNAbin(matrix(names(major_allele_prop), nrow = 1))
  rownames(consensus) <- "Consensus  "

  make_one_plot <- function(y_offset) {
    # The barplot at the top
    par(fig=c(0.206, 1, 0.95 - y_offset, 1 - y_offset), new = TRUE, mar = c(0.5, 4, 1, 0.37))
    barplot(t(as.matrix(barplot_data)), cex.axis = 1, ylab = "PID", col = c('grey', 'red'), border = NA, space = 0, at = c(0, 1))
    # The consensus
    par(fig=c(0.3, 1, 0.94 - y_offset, 0.95 - y_offset), new = TRUE, mar = c(0.1, 0, 0, 2))
    image(consensus, cex.lab = 0.7, show.label = TRUE, yaxt = "n", xaxt = "n", legend = FALSE)
    # The alignment 
    par(fig=c(0.3, 1, 0.5 - y_offset, 0.95 - y_offset), new = TRUE, mar = c(5, 0, 1, 2))
    image(aligned, cex.lab = 0.5, show.label = FALSE, yaxt = "n", xlab = "Alignment position", legend = FALSE)
    # The tree and labels 
    par(fig=c(0, 0.35, 0.5 - y_offset, 0.951 - y_offset), new = TRUE, mar = c(3.9, 4, 0, 0))
    plot(tree, use.edge.length = FALSE, adj = 1, cex = 0.6)
    # The subfigure label
    mtext(title, cex = 2, adj = 0, padj = -1.5)
  }

  # Make plots
  if (spot == "top") {
    make_one_plot(0)
  } else {
    make_one_plot(.5)
  }
}

rps_amp_aligned <- calc_align(rps_amps)
its_amp_aligned <- calc_align(its_amps)

pdf(file = file.path('results', 'alignment_plot.pdf'), width = 8, height = 11)
# png(file = file.path('results', 'alignment_plot.png'), width = 2000, height = 3000, pointsize = 28)
plot_align(rps_amps, rps_amp_aligned, title = expression(italic("rps10")), start = 34)
plot_align(its_amps, its_amp_aligned, title = "ITS1", start = 39, spot = "bottom")
# Add legend
par(fig=c(0.15, 0.5, 0.51, 0.61), new = TRUE, mar = c(0, 0, 0, 0))
legend(0, 0, legend = c('A', 'G', 'C', 'T', '-'), pch = 22, pt.bg = c("red", "yellow", "green", "blue", "black"), 
       pt.cex = 2, bty = "n", xjust = 0.5, yjust = 0.5, 
       horiz = TRUE, xpd = TRUE)
dev.off()
```

```{r}
include_graphics(file.path('results', 'alignment_plot.pdf'))
```

Caption: Multiple sequence alignments of the region predicted to be amplified by each method, not including the primer binding sites. The sequences used represent the subset of species present in both reference databases. The sequences are ordered vertically based on a neighbour-joining tree. Along the top of each alignment there is a barchart representing the proportion of sequences matching the consensus sequence at each alignment position, with gray representing matches and red representing mismatches.

### Calculate distance to nearest non-target

Being able to differentiate different species is key to metabarcoding. 
Therefore, the number of differences between a sequence and the nearest sequence assigned to a different species is useful information.
First I will find which species share identical sequences;

```{r}
calc_identical_sp <- function(aligned, threshold = 1) {
  my_dist <- dist.dna(aligned, model = 'N', as.matrix = TRUE)
  identical_sp <- map(1:nrow(my_dist), function(i) {
    sp_dists <- my_dist[i, -i]
    identical_sp <- unique(names(sp_dists[sp_dists < threshold]))
    identical_sp[identical_sp != rownames(my_dist)[i]]
  })
  names(identical_sp) <- rownames(my_dist)
  unique_identical_sp <- map(unique(names(identical_sp)), function(x) {
    unique(unlist(identical_sp[names(identical_sp) == x]))
  })
  names(unique_identical_sp) <- unique(names(identical_sp))
  out <- lapply(seq_along(unique_identical_sp), function(index) {
  sort(c(names(unique_identical_sp)[index], unique_identical_sp[[index]]))
  })
  out[map_dbl(out, length) > 1 & ! duplicated(out)]
}

rps_ident_sp <- calc_identical_sp(rps_amp_aligned)
its_ident_sp <- calc_identical_sp(its_amp_aligned)
```

and save the results in a table:

```{r}
tibble(locus = rep(c('RPS10', 'ITS1'), c(length(rps_ident_sp), length(its_ident_sp))), 
       identical_seqs = c(map_chr(rps_ident_sp, paste0, collapse = ' ; '), 
                          map_chr(its_ident_sp, paste0, collapse = ' ; '))) %>%
  write_csv(file = file.path('results', 'species_resolution_table.csv'))
```

I will also make a plot of the number of differences between each sequence and their nearest sequence from another species.

```{r fig.width=5, fig.height=5}
calc_nearest_seq_diff <- function(aligned) {
  my_dist <- dist.dna(aligned, model = 'N', as.matrix = TRUE)
  out <- map_dbl(1:nrow(my_dist), function(i) {
    sp_dists <- my_dist[i, -i]
    min(sp_dists[names(sp_dists) != rownames(my_dist)[i]], na.rm = TRUE)
  })
  names(out) <- rownames(my_dist)
  out
}

calc_nearest_seq_diff_dist <- function(aligned, max = 5) {
  diffs <- calc_nearest_seq_diff(aligned)
  uniq_sp_names <- unique(names(diffs))
  diffs <- map_dbl(uniq_sp_names, function(n) {
    max(diffs[names(diffs) == n], na.rm = TRUE)
  })
  names(diffs) <- uniq_sp_names
  out <- map_int(1:max - 1, function(n) {
    sum(diffs == n)
  })
  out[max + 1] <- sum(diffs >= max) 
  names(out) <- c(1:max - 1, paste0(max, '+'))
  out / length(diffs)
}
```

and make a table of that as well:

```{r}
diff_props <- c(calc_nearest_seq_diff_dist(its_amp_aligned), calc_nearest_seq_diff_dist(rps_amp_aligned))
diff_props_data <- tibble(prop = diff_props,
       n = names(diff_props),
       locus = ordered(rep(c('ITS1', 'rps10'), each = 6), levels =  c('rps10', 'ITS1')))
print(diff_props_data)
```

We can plot this in a few ways:

```{r fig.width=5, fig.height=5}
sp_diff_plot <- diff_props_data %>%
  ggplot(aes(x = n, y = prop, fill = locus)) +
  geom_bar(stat = 'identity', position = "dodge2") +
  geom_hline(aes(yintercept = 1), linetype = "dashed") +
  scale_fill_viridis_d(begin = 0.8, end = 0.2) +
  labs(x = 'Base pairs different from most similar species', y = 'Proportion of species', fill = 'Locus') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = c(0.1, 0.8),
        legend.background = element_rect(fill=NA))
sp_diff_plot
ggsave(sp_diff_plot, filename = 'sp_resolution_barchart.pdf', path = file.path('results'), width = 5, height = 5)
```

Figure #: The distribution of the number of base pairs different each species is to the most similar species base on pairwise alignments of predicted amplicons in the Rps10 and ITS1 databases.
Only sequences with unambiguous, species-level taxonomic classifications that contain the entire amplicon are included. Zero differences for a species mean that there is at least one other species that is predicted to have an identical amplicon sequence.

```{r fig.width=2, fig.height=4}
rps_diffs <- calc_nearest_seq_diff(rps_amp_aligned)
its_diffs <- calc_nearest_seq_diff(its_amp_aligned)
diff_boxplot <- tibble(locus = ordered(rep(c('ITS1', 'rps10'), c(length(its_diffs), length(rps_diffs))), levels =  c('rps10', 'ITS1')),
       species = c(names(its_diffs), names(rps_diffs)),
       diff = c(its_diffs, rps_diffs)) %>%
  ggplot(aes(x = locus, y = diff, fill = locus)) +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.8, end = 0.2) +
  guides(fill = FALSE) +
  labs(x = 'Locus', y = 'Base pairs different from most similar species', fill = NULL) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")
diff_boxplot
ggsave(diff_boxplot, filename = 'sp_resolution_boxplot.pdf', path = file.path('results'), width = 2, height = 4)
```

We can combine the two plots like so:

```{r fig.width=7, fig.height=5}
diff_plot_combined <- ggarrange(diff_boxplot, sp_diff_plot,
          labels = c("A", "B"),
          widths = c(0.3, 1),
          ncol = 2, nrow = 1)
diff_plot_combined
ggsave(diff_plot_combined, filename = 'sp_resolution_combined.pdf', path = file.path('results'), width = 7, height = 5)
```

Figure #: The distribution of the number of base pairs different each species is to the most similar species based on pairwise alignments of predicted amplicons in the Rps10 and ITS1 databases.
Only sequences for species present in both databases are included. A) The distribution of differences for each locus. B) Proportion of species with each difference count. Zero differences for a species mean that there is at least one other species that is predicted to have an identical amplicon sequence.

## Software used

```{r}
sessioninfo::session_info()
```

### 2021-09-23 09--simulated_pcr_results.Rmd
make site public

---
title: "Plotting simulated PCR results"
bibliography: '`r sharedbib::bib_path()`'
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
source('style.R')
```

## Prepare

```{r}
library(ape)
library(metacoder)
library(tibble)
```

## Read in results

These results were generated by Felipe Albornoz using Geneious.
They consist of two FASTA files: the reference sequences used for the simulated PCR and the predicted amplicons.

```{r}
ref_seqs <- read.FASTA(file.path("raw_data", "felipes_sim_pcr_results", "all.ref.seqs.rps10.fasta"))
amp_seqs <- read.FASTA(file.path("raw_data", "felipes_sim_pcr_results", "PCR product.fasta"))
```

I will check that all names in `amp_seqs` are in `ref_seqs`

```{r}
if (! all(names(amp_seqs) %in% names(ref_seqs))) {
  stop("All names in `amp_seqs` must be in `ref_seqs`")
}
```

Save location of named seqs in reference

```{r}
amped_indexes <- match(names(amp_seqs), names(ref_seqs))
```

## Clean up headers

The taxonomic information in the FASTA headers need to be cleaned up to be useful for computational analysis and presentation.

```{r}
clean_headers <- function(headers) {
  # remove _F , _sp
  headers <- sub(headers, pattern = "_F$", replacement = "")
  headers <- sub(headers, pattern = "_sp$", replacement = "")

  # Remove existing classifcaitons and only use seq ID
  headers[grepl(headers, pattern = "_Eukaryota;")] <- gsub(headers[grepl(headers, pattern = "_Eukaryota;")],
                                                           pattern = "_Eukaryota;.+$", replacement = "")

  # Remove everything after species name
  matches <- stringr::str_match(headers, "^([a-zA-Z]+_[a-zA-Z]+).*$")[,2]
  headers[! is.na(matches)] <- matches[! is.na(matches)]

  headers <- sub(headers, pattern = "_sp$", replacement = "")
  headers <- sub(headers, pattern = "_x$", replacement = "")

  # Remove underscores
  # headers <- gsub(headers, pattern = "([a-zA-Z]{3,})_", replacement = "\\1 ")

  headers <- sub(headers, pattern = "Peronosppora", replacement = "Peronospora")
  headers <- sub(headers, pattern = "Scleroderma", replacement = "Sclerospora")

  return(headers)
}
cleaned_headers <- clean_headers(names(ref_seqs))
```

Some headers have species names, others have genbank ids. 
I will read in seq ids and names separately and combine classifications.
Since looking up the taxonomy data can be unreliable depending on internet connection, I will cache the results of this in an .rds file.

Seq IDs:

```{r}
seq_id_obj_data_path <- file.path('intermediate_data', 'seq_id_obj.rds')
if (file.exists(seq_id_obj_data_path)) {
  seq_id_obj <- readRDS(seq_id_obj_data_path)
} else {
  seq_id_obj <- lookup_tax_data(cleaned_headers, type = "seq_id")
  saveRDS(seq_id_obj, file = seq_id_obj_data_path)
}
# seq_id_obj <- remove_redundant_names(seq_id_obj)
seq_id_classes <- classifications(seq_id_obj)[names(seq_id_obj$data$query_data)]
```

Seq names:

```{r}
genus_part <- sub(cleaned_headers, pattern = "^([a-zA-Z]+)_.*$",
                  replacement = "\\1")
species_part <- sub(cleaned_headers, pattern = "^[a-zA-Z]+_([a-zA-Z]+)$",
                  replacement = "\\1")
```

I will also save the results of this database search, since it can take a while and it requires manual input when synonyms are encountered, so cannot be included in an Rmd otherwise.

```{r}
taxon_name_obj_data_path <- file.path('intermediate_data', 'taxon_name_obj.rds')
if (file.exists(taxon_name_obj_data_path)) {
  taxon_name_obj <- readRDS(taxon_name_obj_data_path)
} else {
  taxon_name_obj <- lookup_tax_data(genus_part, type = "taxon_name")
  saveRDS(taxon_name_obj, file = taxon_name_obj_data_path)
}
taxon_name_classes <- classifications(taxon_name_obj)[names(taxon_name_obj$data$query_data)]
```

Combine the classifications from the name and ID searches:

```{r}
taxon_name_classes[taxon_name_classes != "unknown taxon"] <- paste0(taxon_name_classes, ";", species_part)[taxon_name_classes != "unknown taxon"]

combined_class <- unname(seq_id_classes)
combined_class[combined_class == "unknown taxon"] <- taxon_name_classes[combined_class == "unknown taxon"]

```

Convert to taxmap format for plotting:

```{r}
obj <- parse_tax_data(combined_class,
                      datasets = list(original_names = names(ref_seqs)),
                      mappings = c("{{index}}" = "{{index}}"))
```

## Filter out ambigous taxa and atrifacts

This will make the tree cleaner and easier to interpret.

```{r}
cleaned_obj <- obj %>%
  filter_taxa(is_leaf, taxon_names %in% c("Glomus", "Phytophthora"), invert = TRUE) %>%
  filter_taxa(grepl(taxon_names, pattern = 'unclassified', ignore.case = TRUE), invert = TRUE) %>%
  filter_taxa(grepl(taxon_names, pattern = 'endosymbiont', ignore.case = TRUE), invert = TRUE) %>%
  filter_ambiguous_taxa() %>%
  filter_taxa(grepl(taxon_names, pattern =  "^[a-zA-Z ]+$"))

cleaned_obj$data$tax_data <- tibble(taxon_id = names(cleaned_obj$data$tax_data), 
                                    input = cleaned_obj$data$tax_data)
```

## Plot what was amplified

I will find which portions of the taxonomy have all of their leaves amplified ...

```{r}
cleaned_obj$mutate_obs("is_amplified",
               unlist(leaves_apply(cleaned_obj,
                                    function(x) length(x) > 0 && all(x %in% names(amp_seqs)),
                                    value = "original_names")))
amplified_leafs <- cleaned_obj$data$original_names[cleaned_obj$data$original_names %in% names(amp_seqs)]
cleaned_obj$data$is_amplified[names(amplified_leafs)] <- TRUE
```

... and plot those taxa in green:

```{r}
set.seed(5)
cleaned_obj %>% 
  filter_taxa(!is_internode) %>%
  remove_redundant_names() %>%
  heat_tree(node_color = ifelse(is_amplified, "green", "grey"),
            node_color_axis_label = "PID to closest other species",
            node_size = n_obs, 
            node_size_range = c(0.005, 0.025),
            node_label_size_range = c(0.012, 0.018),
            # node_label_size_trans = "log10",
            node_size_axis_label = "Number of sequences",
            node_label = Hmisc::capitalize(taxon_names),
            layout = "da", initial_layout = "re",
            background_color = '#FFFFFF',
            # overlap_avoidance = 2,
            output_file = file.path("results", "rps10_simulated_pcr.pdf"))
```

### 2021-09-23 index.Rmd
make site public

---
title: "Introduction"
---

This analysis evaluates the effectiveness of the rps10 locus and the primers we designed for use in Oomycete metabarcoding compared to the current ITS1-based method using ITS6 and ITS7.
The criteria for an ideal barcode and assocaited primers for metabarcoding are:

1) The primers can amplify the gene in all of the target group of organisms.
2) The primers do not amplify the gene in any organisms not in the target group.
3) The sequence amplified can differentiate all members of the target group.
4) The reference database contains sequences for all members of the target group.

We have attempted to evaluate each of these criteria to compared methods using RPS10 and ITS1.
This website is the output of the analysis we used to create the results reported in [insert publication].

### 2021-09-23 style.R

@@ -1,0 +1,2 @@
+knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 10)
+# options(width = 100)

### 2021-09-23 tools.R

@@ -1,0 +1,157 @@
+
+
+
+#' @keywords internal
+parse_ft <- function(text) {
+  text <- gsub(text, pattern = "\n\t\t\t", replacement = "\t", fixed = TRUE)
+  parts <- strsplit(text, "\n\n", fixed = TRUE)[[1]]
+  part_data <- lapply(parts, function(x) {
+    first_line <- sub(x, pattern = "\\n.+$", replacement = "")
+    acc <- stringr::str_match(first_line, pattern = "\\|(.+)\\|")[,2]
+    the_rest <- sub(x, pattern = "^>.+?\\n", replacement = "")
+    # replace extra \t with , when there are multiple features
+    lines <- strsplit(the_rest, "\n")[[1]]
+    lines <- purrr::map2(stringr::str_locate_all(lines, "\t"), lines, function(matches, line) {
+      if (nrow(matches) > 4) {
+        for (i in matches[3:(nrow(matches) - 2), 1]) {
+          substr(line, i, i) <- ","
+        }
+      }
+      return(line)
+    })
+    the_rest <- paste0(lines, collapse = "\n")
+    output <- readr::read_tsv(paste0(the_rest, "\n"), col_names = c("start", "end", "feature", "type", "name"), col_types = "ccccc")
+    output <- tibble::as_tibble(cbind(list(acc = acc), output, stringsAsFactors = FALSE))
+  })
+  output <- dplyr::bind_rows(part_data)
+  output$complete <- ifelse(startsWith(as.character(output$start), "<") | startsWith(as.character(output$end), ">"), FALSE, TRUE)
+  output$start <- as.integer(gsub(output$start, pattern = "<", replacement = ""))
+  output$end <- as.integer(gsub(output$end, pattern = ">", replacement = ""))
+  return(output)
+}
+
+#' @keywords internal
+parse_seqs <- function(text) {
+  xml <- xml2::read_xml(text)
+  tibble::tibble(acc = xml2::xml_text(xml2::xml_find_all(xml, "//TSeq_accver")),
+                 seq = xml2::xml_text(xml2::xml_find_all(xml, "//TSeq_sequence")),
+                 header = xml2::xml_text(xml2::xml_find_all(xml, "//TSeq_defline")),
+                 length = xml2::xml_text(xml2::xml_find_all(xml, "//TSeq_length")))
+}
+
+#' Lookup gene sequences from NCBI
+#'
+#' Look for sequences of a particular gene for a list of species/isolates/genes from the Genbank
[…]

### 2021-09-23 index.Rmd
fix website

@@
----
-title: "Introduction"
----
+## Analysis code for "A new oomycete metabarcoding method using the rps10 gene"
+
+Authors: Zachary S.
+L.
+Foster, Felipe E Albornoz, Valerie J Fieland, Meredith M Larsen, Frank Andrew Jones, Brett M Tyler, Hai D.
+T.
+Nguyen, Treena I Burgess, Carolyn Riddell, Hermann Voglmayr, Frank N Martin, Niklaus J Grunwald
+
+bioRxiv 2021.09.22.460084; doi: https://doi.org/10.1101/2021.09.22.460084
 
 This analysis evaluates the effectiveness of the rps10 locus and the primers we designed for use in Oomycete metabarcoding compared to the current ITS1-based method using ITS6 and ITS7.
-The criteria for an ideal barcode and assocaited primers for metabarcoding are:
-
-1) The primers can amplify the gene in all of the target group of organisms.
-2) The primers do not amplify the gene in any organisms not in the target group.
-3) The sequence amplified can differentiate all members of the target group.
-4) The reference database contains sequences for all members of the target group.
-
-We have attempted to evaluate each of these criteria to compared methods using RPS10 and ITS1.
-This website is the output of the analysis we used to create the results reported in [insert publication].

### 2022-03-23 07--nontarget_amplification.Rmd
working on reviewer request

@@
 
+## Look for overlap in oomycetes detected
+
+A reviewer requested that we look to see if any oomycetes are detected by ITS1 that are not detected by rps10.
+
+```{r}
+obj <- parse_tax_data(abundance_asv, class_cols = 'taxonomy', class_sep = ';',
+                      class_regex = '^(.+)--(.+)--(.+)$',
+                      class_key = c(taxon = 'taxon_name', boot = 'info', rank = 'taxon_rank'))
+obj$data$class_data <- NULL
+
+# Just look at oomycetes
+obj <- filter_taxa(obj, taxon_names == "Oomycetes", supertaxa = FALSE, subtaxa = TRUE)
+
+# Just look at species data
+obj <- filter_taxa(obj, taxon_ranks == "Species", supertaxa = FALSE)
+
+# Sum data by taxon
+obj$data$tax_abund <- calc_taxon_abund(obj, data = "tax_data", cols = metadata$sample_id, groups = metadata$locus)
+
+# Color by locus
+obj$data$tax_abund$locus <- "Both"
+obj$data$tax_abund$locus[obj$data$tax_abund$rps10 >= minimum_read_count & obj$data$tax_abund$ITS < minimum_read_count] <- 'Rps10'
+obj$data$tax_abund$locus[obj$data$tax_abund$rps10 < minimum_read_count & obj$data$tax_abund$ITS >= minimum_read_count] <- 'ITS1'
+table(obj$data$tax_abund$locus)
+
+# Plot
+obj %>%
+  heat_tree(node_label = taxon_names,
+            node_size = n_obs)
+
+```
+
 ## Software used

### 2022-03-23 07--nontarget_amplification.Rmd
more filtering which species are used for the comparison

--- 2021-09-23 07--nontarget_amplification.Rmd
@@
 
+## Look for overlap in oomycetes detected
+
+A reviewer requested that we look to see if any oomycetes are detected by ITS1 that are not detected by rps10.
+
+```{r}
+obj <- parse_tax_data(abundance_asv, class_cols = 'taxonomy', class_sep = ';',
+                      class_regex = '^(.+)--(.+)--(.+)$',
+                      class_key = c(taxon = 'taxon_name', boot = 'info', rank = 'taxon_rank'))
+
+# Remove ASVs without confident taxonomic assignments
+is_confident_species <- map_lgl(split(obj$data$class_data, obj$data$class_data$input_index), function(x) {
+  as.numeric(x$boot[x$rank == "Species"]) >= 50
+})
+ref_pid <- as.numeric(str_match(obj$data$tax_data$taxonomy, pattern = '--([0-9.]+)--ASV$')[, 2])
+obj <- filter_obs(obj, data = "tax_data", is_confident_species & ref_pid > 98, drop_taxa = TRUE)
+obj$data$class_data <- NULL
+
+# Remove taxa not in both databases
+its_ref <- read_fasta(file.path('intermediate_data', 'reference_databases', 'its1_reference_db.fa'))
+rps10_ref <- read_fasta(file.path('intermediate_data', 'reference_databases', 'rps10_reference_db.fa'))
+in_both_db <- map_lgl(taxon_names(obj), function(n) {
+  any(grepl(names(its_ref), pattern = n, ignore.case = TRUE)) & any(grepl(names(rps10_ref), pattern = n, ignore.case = TRUE))
+})
+obj <- filter_taxa(obj, in_both_db, supertaxa = TRUE)
+
+# Just look at oomycetes
+obj <- filter_taxa(obj, taxon_names == "Oomycetes", supertaxa = FALSE, subtaxa = TRUE)
+
+# Just look at species data
+obj <- filter_taxa(obj, taxon_ranks == "Species", supertaxa = FALSE)
+
+# Sum data by taxon
+obj$data$tax_abund <- calc_taxon_abund(obj, data = "tax_data", cols = metadata$sample_id, groups = metadata$locus)
+obj$data$tax_abund$name <- taxon_names(obj)[obj$data$tax_abund$taxon_id]
+
+# Summarize by locus
+obj$data$tax_abund$locus <- "Both"
+obj$data$tax_abund$locus[obj$data$tax_abund$rps10 >= minimum_read_count & obj$data$tax_abund$ITS < minimum_read_count] <- 'Rps10'
+obj$data$tax_abund$locus[obj$data$tax_abund$rps10 < minimum_read_count & obj$data$tax_abund$ITS >= minimum_read_count] <- 'ITS1'
+table(obj$data$tax_abund$locus)
+```
+
 ## Software used

### 2022-03-24 06--mock_community.Rmd
create plot of mock comm read prop for reviewer

@@
 
-and I can add which reference sequence each asv was assigned to and its PID as seperate columns for easy use afterwards.
+and I can add which reference sequence each asv was assigned to and its PID as separate columns for easy use afterwards.
 
@@
 rps10_align_data <- align_mock_comm_seqs(asv_data, 'rps10', rps10_ref_data)
-as.data.frame(filter_to_top_mc_ref_hits(rps10_align_data))
+(rps10_best_mc_hits <- as.data.frame(filter_to_top_mc_ref_hits(rps10_align_data)))
 ```
@@
 its_align_data <- align_mock_comm_seqs(asv_data, 'ITS', its_ref_data)
-as.data.frame(filter_to_top_mc_ref_hits(its_align_data))
+(its_best_mc_hits <- as.data.frame(filter_to_top_mc_ref_hits(its_align_data)))
+```
+
+We can plot this to check how even the read depth is for each locus:
+
+```{r}
+rps10_best_mc_hits$abund <- rps10_best_mc_hits$abund / sum(rps10_best_mc_hits$abund)
+its_best_mc_hits$abund <- its_best_mc_hits$abund / sum(its_best_mc_hits$abund)
+plot_data <- rbind(mutate(rps10_best_mc_hits, locus = "Rps10"), mutate(its_best_mc_hits, locus = "ITS1"))
+mc_read_prop_plot <- plot_data %>%
+  ggplot(aes(x = locus, y = abund)) +
+  geom_boxplot() +
+  geom_hline(yintercept=1/24, linetype="dashed") +
+  labs(x = "Locus", y = "Read Proportion")
+mc_read_prop_plot
+ggsave(mc_read_prop_plot, filename = 'read_proportion_of_mock_community.pdf', path = file.path('results'), width = 5, height = 5)
+```
+
+Standard deviation:
+
+```{r}
+sd(rps10_best_mc_hits$abund)
+sd(its_best_mc_hits$abund)
+```
+
+We can also look at if the proportion of sequences in correlated
+
+```{r}
+plot_data %>%
+  pivot_wider(names_from = locus, values_from = abund) %>%
+  ggplot(aes(x = Rps10, y = ITS1)) +
+  geom_point()
 ```

### 2022-03-24 06--mock_community.Rmd
imporved read proporion plot for reviewer

@@
 library(Biostrings)
+library(ggpubr)
 ```
@@
   geom_hline(yintercept=1/24, linetype="dashed") +
-  labs(x = "Locus", y = "Read Proportion")
+  labs(x = "Locus", y = "Read Proportion") +
+  theme_minimal() 
 mc_read_prop_plot
-ggsave(mc_read_prop_plot, filename = 'read_proportion_of_mock_community.pdf', path = file.path('results'), width = 5, height = 5)
 ```
@@
 ```{r}
-plot_data %>%
+plot_data$obligate <- plot_data$species %in% c("Peronospora effusa", "Peronospora schachtii", "Plasmopara halstedii", "Plasmopara obducens", "Pseudoperonospora cubensis")
+plot_data$low_dna <- plot_data$species %in% c("Phytophthora pluvialis", "Saprolegnia diclina")
+plot_data$dna_type <- "Normal"
+plot_data$dna_type[plot_data$obligate] <- "Obligate"
+plot_data$dna_type[plot_data$low_dna] <- "Less DNA"
+plot_data$species[plot_data$species == "Pythium citrinum"] <- "Phytopythium citrinum"
+read_prop_corr_plot <- plot_data %>%
+  select(species, abund, locus, obligate, low_dna, olbigate_or_low_dna, dna_type) %>%
   pivot_wider(names_from = locus, values_from = abund) %>%
-  ggplot(aes(x = Rps10, y = ITS1)) +
-  geom_point()
+  ggplot(aes(x = Rps10, y = ITS1, color = dna_type)) +
+  geom_point(alpha= 0.5) +
+  coord_fixed(xlim = c(0, 0.3), ylim = c(0, 0.3)) +
+  labs(x = "Rps10 read proportion", y = "ITS1 read proportion", color = "DNA type") +
+  theme_minimal() 
+read_prop_corr_plot
+```
+
+Lets combine these two plots since they both describe the same data
+
+```{r}
+mc_plot_combined <- ggarrange(mc_read_prop_plot, read_prop_corr_plot,
+                              labels = c("A", "B"),
+                              widths = c(.55, 1),
+                              ncol = 2, nrow = 1)
+mc_plot_combined
+ggsave(mc_plot_combined, filename = 'read_proportion_of_mock_community.pdf', path = file.path('results'), width = 10, height = 5)
 ```

### 2022-03-24 publication/reviewer_response.Rmd
revision almost done

# Reviewer 1

> This research will be very useful for people working on metabarcoding projects across many fields. The RPS10 loci appears to be well suited for barcoding oomycetes using Miseq sequencing technology and the creation of a website for the community to contribute reference sequences is a good contribution that will give ongoing value to the research community. I have made some suggestions for minor revisions below. As an aside, the website isn't under review here but could I also suggest that you might include some bioinformatic workflows there as well. This would useful for reproducibility for others in the future.

Thank you for your review and constructive feedback!
We are considering ways to improve the website and would like to have a few more tools there in the future.

> 55 : “Evaluated” might be a better word than “Identified” here

Done.

> 57 : Change “amplify” to “amplify this region from”

Done.

> 79: Change “often” to “can”

Done.

> 96: “Eucalypt” should be changed to “Eucalyptus”

Done.

> 109: Change “fungi and bacteria” to “fungi or bacteria”

Done.

> 110: Could you please expand a little about what you mean about activity of oomycetes.

The statement was referring to some types of oomycetes that must be baited during certain seasons when they are sporulating.
However, since that does not apply to all oomycetes and is vague without more explanation, we have removed that portion of the sentence.

> 176: Change “amplification” to “PCR”

Done.

> 185 “Efficacy” might read better than “usefulness”

Done.

> 185 – 188 : the way this is phrased it sound like you are comparing rps10 primers ability to amplify the ITS1 region with the Sapkota primers which they don’t amplify. Could you re word it please to be a clearer.

Done.

> 205: It would be useful and informative if you included the alignment as a supplementary file

!!!!TODO!!!! (Waiting on Frank to get back to us)

> 211: “All known oomycetes” implies you checked every oomycete ever described

Good point. We meant that the goal was for the primers to amplify all oomycetes, but I see how that could be misinterpreted.
We have modified the text.

> 214: Which cutoffs did you use for the parameters such as GC content, hairpins etc when selecting primers?

Rigid cutoffs were not used, but rather people experienced with primer design compared the OligoAnalyzer results to their previous experiences.

> 239: How was the DNA quantified?

We used a Qubit. This information has been added.

> 321 – 322: This first sentence is not required because the information is repeated in more detail below. Please just add the references for DADA2 and Cutadapt where approcpriate below instead.

Done.

> 478 – 494: Were there any oomycetes amplified/identified with the ITS1 primer set that were not identified with the rps10 primers?

Unfortunately the data from the environmental samples cannot be used to answer this question confidently for two reasons:

* The content of the reference databases differs for the two loci, so even if both methods amplified the same organism, the closest match in the reference database of each locus could be to different taxa. Since many ASVs in the environmental samples have no close match in either database (only 16% of oomycete ASVs from environmental samples had a match >95% identical), this error is likely to occur often.
* The limited taxonomic resolution of ITS1, and rps10 to a lesser extent, means that even if both databases have a reference sequence perfectly matching an ASV, there could be other taxa in that database with the same sequence, so which taxon name gets assigned to the ASV is random.

This is why the mock community was useful; in that case we knew all the organisms sequenced had a perfect match in both databases.

> 516 Please change “are generally more diverged from each other” to “are generally more divergent”

Done.

> 540 - 541: Please change “showed” to “predicted” also please change “non-target taxa” to “all non-target taxa tested”

Thanks for pointing that out. Done.

# Reviewer 2

> This study describes the design of a valuable new barcoding locus/assay developed from a database of mtDNA sequence data. Its utility was then compared carefully against an ITS rDNA locus. The methods were used to metabarcode a mixed mock community and on eDNA samples from natural environments. The paper is a well-written account of a valuable breakthrough in oomycete barcoding and should be published.  A key benefit of the rps10 locus compared to the ITS loci primers seems to me the ability to amplify products beyond the Phytophthora and downy mildew genera to include other oomycetes.  I do, however have a few comments suggested changes that I feel would add to value of the publication, in particular in relation to the read ratio in the mock community.  I also feel it may be helpful to detail which of the reverse primers has specificity to which group of oomycetes. Could this be used to give future researchers an option to pick and choose the combination of rps10 primers to focus on the group of oomycetes of particular interest?

> As an aside: note that the primers ITS6 and 7 were never designed to be, or described as, an oomycete-specific assay by Cooke et al (2000) but were simply modifications of the existing White et al (1990) fungal ITS primers to allow more robust amplification of oomycete samples in the laboratory. They were taken up and tested by other authors as ‘oomycete specific’. It was not entirely surprising they were not effective as the discriminating bases are often not at that the 3’ end that would have offered greater specificity.  I say this not to criticise the rps10 assay which looks very powerful, but the point could be made that an ITS-based assay that was as carefully designed as the rps10 one could also offer value for oomycete barcoding.  The assay of Sapkota and Nicolaisen (2015) was not ‘designed’ but tested ‘in hope’ and is maybe not a fair comparison. Having said that, there were no better oomycete ITS assays to be tested so this papers detailed investigation is valuable and will hopefully prevent others using these ITS primers as oomycete-specific in the future. While a comparison may have been with the Scibetta et al (2012) ITS primers which are highly specific to known downy mildew and Phytophthora I understand that these do not amplify other oomycetes and would not have been equivalent. The key advantages of rps10 – oomycete specificity and improved taxonomic resolution - are clear.

These are excellent points.
We have added a sentence to explain the history of ITS6 and ITS7.

> What happens with Nothophytophthora? I could not see them on Figure 2. There is no reason to suggest they would not amplify but perhaps they should be mentioned somewhere?

The data used to make Figure 2 was collected during the initial design of the primers.
At the time Nothophytophthora had just been described and we were not aware of any publicly available rps10 sequences for Nothophytophthora that included the primer binding sites.
As far as we can tell, the rps10 sequences for Nothophytophthora on NCBI currently available do not have the primer binding sites so we cannot predict whether or not Nothophytophthora would be amplified by this method.
Nothophytophthora is one of many genera not represented by currently available rsp10 sequences.
We have added a sentence to the methods (around line 228) using Nothophytophthora as an example for genera we dont have data for.

> The web-site https://nam04.safelinks.protection.outlook.com/?url=http%3A%2F%2Fwww.oomycetedb.org%2F&amp;data=04%7C01%7Cniklaus.grunwald%40oregonstate.edu%7C04b24f72622c4cf9996308da05c111ce%7Cce6d05e13c5e4d6287a84c4a2713c113%7C0%7C0%7C637828624036753211%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000&amp;sdata=%2FlQvEfQI7iNfh7VRxfzITmsi4h2dliKEmNs2dGUVj80%3D&amp;reserved=0 is a valuable resource.

Thank you!

> L145 “due to it robust statistical” needs an extra word or ‘its’

Done.

> L153 “currently available ITS1” I would say “some currently available ITS1” as there are ones that do not such as the unmentioned Scibetta et al primers. I feel reference should be made to the source paper (Scibetta et al 2012) which described the downy mildew and Phytophthora-specific primers employed in several key studies such as those of Landa and Riddell.
> Scibetta, S.; Schena, L.; Chimento, A.; Cacciola, S.O.; Cooke, D.E. A molecular method to assess Phytophthora diversity in environmental samples. J. Microbiol. Methods 2012, 88, 356–368, doi:10.1016/j.mimet.2011.12.012.

Good observation.
A reference to the Scibetta paper has been added.

> L157 “as little as 5.3% of” simpler just to say “only 5.3% of ”

Done.

> L185 suggest “power” or “specificity“ rather than ”usefulness” or leave and just say they were compared.

Another reviewer suggested "efficacy" and we used that, but thanks for the suggestion!

> L294 should specify the version of Illumina chemistry employed.  v2 or v3 and how many cycles?  This is important.

Done.

> L471 onwards – for the mock community it would also be interesting to know the read numbers for each species. Were all reads balanced in 1/24 ratio? This would give a measure of the uniformity of amplification efficiency of both assays.  For example did the unknown proportion of downy mildew vs plant DNA result in fewer DM species reads? How did ITS and mtDNA compare? The more uniform amplicon length of rps10 compared to ITS1 should theoretically offer an advantage as it may minimise competition in which shorter amplicons are more abundant than longer ones due to more efficient amplification. Was this the case? I understand the assays are not intended to be quantitative but it would be useful to know if differences in read number broadly align with initial DNA concentration.

Good idea. This information has been added.

> SE Comment: Is there a spot you could include the importance to Phytobiome research, maybe early in the introduction. It is just nice to see the word phytobiome/s used.

Done.

### 2022-03-28 09--simulated_pcr_results.Rmd
added primer binding site analysis requested by reviewer

@@
 library(tibble)
+library(stringr)
+library(readr)
+library(dplyr)
+library(furrr)
+library(Biostrings)
 ```
 
-## Read in results
+## Plot Felipe's analysis results
+
+### Read in results
 
@@
 
-## Clean up headers
+### Clean up headers
 
@@
 
-## Filter out ambigous taxa and atrifacts
+### Filter out ambigous taxa and atrifacts
 
@@
 
-## Plot what was amplified
+### Plot what was amplified
 
@@
 ```
+
+## Which primers amplify which oomyctes
+
+A reviewer thought it would be useful to know which of the variations of the primer amplified which organisms, with the hope that researchers could use subsets of the primers to target specific groups.
+
+### Extract all oomycete from reference databases
+
+```{r}
+rps_ref <- read_fasta(file.path('intermediate_data', 'reference_databases', 'rps10_reference_db.fa'))
+rps_ref <- rps_ref[grepl(names(rps_ref), pattern = 'Oomycetes')]
+```
+
+Remove spaces from sequences (not sure why there would be spaces).
+
+```{r}
+rps_ref <- trimws(gsub(rps_ref, pattern = ' ', replacement = ''))
+```
+
+Replace names with just species names
+
+```{r}
+names(rps_ref) <- str_match(names(rps_ref), ';([a-zA-Z0-9_\\-\'". ]+);oodb_.+;$')[,2]
+```
+
+Replace underscores with spaces
+
+```{r}
+names(rps_ref) <- gsub(names(rps_ref), pattern = '_', replacement = ' ')
+```
+
+Remove species with ambiguous names or numbers
+
+```{r}
+is_ambiguous <- function(n) {
+  grepl(n, pattern = ' sp$') |
+    grepl(n, pattern = ' spp$') | 
+    grepl(n, pattern = ' spp.') |
+    grepl(n, pattern = ' sp.') |
+    grepl(n, pattern = ' aff. ')  | 
+    grepl(n, pattern = ' cf.') |
+    grepl(n, pattern = '[0-9]+') 
+}
+rps_ref <- rps_ref[! is_ambiguous(names(rps_ref))]
+```
+
+Remove information below species
+
+```{r}
+names(rps_ref) <- sub(names(rps_ref), pattern = ' var\\.?.+$', replacement = '')
+names(rps_ref) <- sub(names(rps_ref), pattern = ' subsp\\.?.+$', replacement = '')
+```
+
+Now the names should be cleaned up enough:
+
+```{r}
+unique(names(rps_ref))
+```
+
+### Extract predicted amplicons
+
+Not all reference sequences will have the entire region.
+Some will have the region but not the primer sites, so these must be removed.
+
+```{r}
+primer_data <- read_csv(file.path("raw_data", "primer_data.csv"))
+primer_data <- filter(primer_data, primer_id != 'rps10_f-F', locus == "rps10")
+forward <- primer_data$sequence[primer_data$direction == "Forward"]
+reverse <- primer_data$sequence[primer_data$direction == "Reverse"]
+```
+
+The `matchProbePair` function from the `Biostrings` package can extract a region predicted to be amplified by primers, but for some reason it does not allow for ambiguity codes, even though it is a trivial change to make it do so.
+Below is a modified version of the function that does allow ambiguity codes by providing the `fixed` parameter and passing it to the `matchPattern` calls.
+
+```{r}
+matchProbePair <- function(Fprobe, Rprobe, subject, algorithm="auto",
+             logfile=NULL, verbose=FALSE, fixed = FALSE)
+    {
+        ## This won't copy the data if Fprobe and Rprobe are already DNAString objects
+        F <- DNAString(Fprobe)
+        R <- DNAString(Rprobe)
+
+        ## F and R hits on the + strand
+        Fp_hits <- start(matchPattern(F, subject, algorithm=algorithm, fixed = fixed))
+        Rp_hits <- start(matchPattern(R, subject, algorithm=algorithm, fixed = fixed))
+
+        ## F and R hits on the - strand
+        Fm_hits <- end(matchPattern(reverseComplement(F), subject, algorithm=algorithm, fixed = fixed))
+        Rm_hits <- end(matchPattern(reverseComplement(R), subject, algorithm=algorithm, fixed = fixed))
+
+        if (verbose) {
+            cat("Fp_hits:", Fp_hits, "  Rp_hits:", Rp_hits,
+                "  Fm_hits:", Fm_hits, "  Rm_hits:", Rm_hits, "\n")
+        }
+
+        matches0 <- Biostrings:::reduceProbePairMatches(c(Fp_hits, Rp_hits), c(Fm_hits, Rm_hits))
+        ans <- Views(subject, start=matches0$start, end=matches0$end)
+
+        if (!is.null(logfile)) {
+            nFp <- length(Fp_hits)
+            nRp <- length(Rp_hits)
+            nFm <- length(Fm_hits)
+            nRm <- length(Rm_hits)
+            nmatches0 <- length(ans)
+            ## cat("", ..., sep="\t") is a trick to get an extra tab
+            cat("", nFp, nRp, nFm, nRm, nmatches0, file=logfile, sep="\t")
+        }
+        ans
+}
+```
+
+The code below extracts predicted amplicons using primer matches, removes the primers, and aligns the trimmed amplicons to the sequences that did not have primer matches to try to find other sequences in the reference databases that do not have primer sequences but are full length matches to those that do.
+
+```{r}
+get_amplicon_chr <- function(seq) {
+  map2_chr(seq@ranges@start, seq@ranges@width, function(s, w) {
+    as.character(seq@subject[seq(from = s, length.out = w)])
+  })
+}
+
+calc_amplicons <- function(seqs, forward, reverse, min_coverage = 0.9, ...) {
+  # Get simulated amplicons using primers
+  full_amps <- future_map(seqs, function(s) {
+    get_amplicon_chr(matchProbePair(DNAString(s), Fprobe = forward, Rprobe = reverse))
+  })
+
+  # Check for multiple possible amplicons per input
+  if (any(map_int(full_amps, length) > 1)) {
+    stop('Some inputs have more than one amplicons.')
+  }
+  full_amps <- unlist(full_amps[map_int(full_amps, length) == 1])
+
+  return(full_amps)
+}
+
+rps_amps <- calc_amplicons(rps_ref, 
+                           forward = primer_data$sequence[primer_data$direction == 'Forward' & primer_data$locus == 'rps10'],
+                           reverse = primer_data$sequence[primer_data$direction == 'Reverse' & primer_data$locus == 'rps10'])
+```
+
+The amplicon could not be extracted from many sequences, probably because it was incomplete.
+Here is the proportion of reference sequences with full length amplicons extracted:
+
+```{r}
+length(rps_ref)
+length(rps_amps)
+```
+
+Lets remove any sequences that are duplicates (same sequence and same species name):
+
+```{r}
+rps_amps <- rps_amps[! duplicated(paste0(names(rps_amps), rps_amps))]
+length(rps_amps)
+```
+
+I noticed when debugging an issue further on in the analysis that at least one sequence is reverse complimented relative to the other sequences.
+I will fix that here:
+
+```{r}
+to_rev_comp <- ! grepl(rps_amps, pattern = "^GTTGGTT")
+rps_amps[to_rev_comp] <- rev_comp(rps_amps[to_rev_comp])
+```
+
+### List primer binding site sequences for each species
+
+Cut out the primer binding region for each primer:
+
+```{r}
+forward_bindings <- substr(rps_amps, start = 0, stop = nchar(forward))
+reverse_bindings <- substr(rps_amps, start = nchar(rps_amps) - nchar(reverse) + 1, stop = nchar(rps_amps)) %>%
+  rev_comp()
+```
+
+check that the primer binding sites are the right sequences for the primers:
+
+```{r}
+stopifnot(all(grepl(forward_bindings, pattern = "^GTTGGTTAGAG[CT]A[AG]AAGACT$")))
+stopifnot(all(grepl(reverse_bindings, pattern = "^AT[AG][CT][CT]TAGAAAGA[CT]T[CT]GAACT$")))
+```
+
+We can put this information in a table
+
+```{r}
+binding_data <- tibble(species = names(forward_bindings),
+                       genus = sub(species, pattern = ' .+$', replacement = ''),
+                       forward = forward_bindings,
+                       reverse = reverse_bindings,
+                       amplicon = rps_amps)
+```
+
+and save it for later reference:
+
+```{r}
+write_csv(binding_data, file = file.path('results', 'primer_binding_site_sequences.csv'))
+```
+
+### Check if primers correspond to genus
+
+Some primers are much more common than others:
+
+```{r}
+binding_data$primer_pair <- paste(forward_bindings, "+", reverse_bindings)
+table(binding_data$forward)
+table(binding_data$reverse)
+table(binding_data$primer_pair)
+```
+
+They don't seems to correspond the genus differences generally, but perhaps a few groups have an unique primer pair:
+
+```{r}
+t(table(binding_data[, c("genus", "forward")]))
+t(table(binding_data[, c("genus", "reverse")]))
+t(table(binding_data[, c("genus", "primer_pair")]))
+```

### 2022-03-28 09--simulated_pcr_results.Rmd
saved output table for genus counts for primer binding sites

@@
 ```{r}
-t(table(binding_data[, c("genus", "forward")]))
-t(table(binding_data[, c("genus", "reverse")]))
-t(table(binding_data[, c("genus", "primer_pair")]))
+(genus_bindings_forward <- t(table(binding_data[, c("genus", "forward")])))
+(genus_bindings_reverse <- t(table(binding_data[, c("genus", "reverse")])))
+(genus_bindings_pair <- t(table(binding_data[, c("genus", "primer_pair")])))
 ```
+
+Lets save all that information in a table for future reference:
+
+```{r}
+genus_bindings_data <- rbind(
+  cbind(type = 'forward', genus_bindings_forward),
+  cbind(type = 'reverse', genus_bindings_reverse),
+  cbind(type = 'pair', genus_bindings_pair)
+)
+genus_bindings_data <- cbind(primer = rownames(genus_bindings_data), genus_bindings_data)
+rownames(genus_bindings_data) <- NULL
+write_csv(as.data.frame(genus_bindings_data),
+          file = file.path('results', 'genus_primer_binding_site_counts.csv'))
+```

### 2022-03-30 publication/reviewer_response.Rmd
update reveiew response from thumb drive

@@
+---
+output:
+  word_document: default
+---
 # Reviewer 1
 
-> This research will be very useful for people working on metabarcoding projects across many fields.
+*This research will be very useful for people working on metabarcoding projects across many fields.
 The RPS10 loci appears to be well suited for barcoding oomycetes using Miseq sequencing technology and the creation of a website for the community to contribute reference sequences is a good contribution that will give ongoing value to the research community.
@@
 As an aside, the website isn't under review here but could I also suggest that you might include some bioinformatic workflows there as well.
-This would useful for reproducibility for others in the future.
+This would useful for reproducibility for others in the future.*
 
 Thank you for your review and constructive feedback!
-We are considering ways to improve the website and would like to have a few more tools there in the future.
+We are considering ways to improve the website and would like to have more tools available there in the future.
 
-> 55 : “Evaluated” might be a better word than “Identified” here
+*55 : “Evaluated” might be a better word than “Identified” here*
 
@@
 
-> 57 : Change “amplify” to “amplify this region from”
+*57 : Change “amplify” to “amplify this region from”*
 
@@
 
-> 79: Change “often” to “can”
+*79: Change “often” to “can”*
 
@@
 
-> 96: “Eucalypt” should be changed to “Eucalyptus”
+*96: “Eucalypt” should be changed to “Eucalyptus”*
 
@@
 
-> 109: Change “fungi and bacteria” to “fungi or bacteria”
+*109: Change “fungi and bacteria” to “fungi or bacteria”*
 
@@
 
-> 110: Could you please expand a little about what you mean about activity of oomycetes.
+*110: Could you please expand a little about what you mean about activity of oomycetes.*
 
 The statement was referring to some types of oomycetes that must be baited during certain seasons when they are sporulating.
-However, since that does not apply to all oomycetes and is vague without more explanation, we have removed that portion of the sentence.
+For example, Phytophthora species in streams are difficult to bait some times of the year (Reeser 2011, Figure 2).
+However, since that does not apply to all oomycetes and is too vague without more explanation, we have removed that portion of the sentence.
 
-> 176: Change “amplification” to “PCR”
+Reeser, P.W., Sutton, W., Hansen, E.M., Remigi, P. and Adams, G.C., 2011. Phytophthora species in forest streams in Oregon and Alaska. Mycologia, 103(1), pp.22-35.
+
+!!!!CHECK WITH NIK!!!!
+
+*176: Change “amplification” to “PCR”*
 
@@
 
-> 185 “Efficacy” might read better than “usefulness”
+*185 “Efficacy” might read better than “usefulness”*
 
@@
 
-> 185 – 188 : the way this is phrased it sound like you are comparing rps10 primers ability to amplify the ITS1 region with the Sapkota primers which they don’t amplify.
-Could you re word it please to be a clearer.
+*185 – 188 : the way this is phrased it sound like you are comparing rps10 primers ability to amplify the ITS1 region with the Sapkota primers which they don’t amplify.
+Could you re word it please to be a clearer.*
 
@@
 
-> 205: It would be useful and informative if you included the alignment as a supplementary file
+*205: It would be useful and informative if you included the alignment as a supplementary file*
 
-!!!!TODO!!!! (Waiting on Frank to get back to us)
+!!!!TODO!!!! (Waiting for Frank to get back to us with the alignment)
 
-> 211: “All known oomycetes” implies you checked every oomycete ever described
+*211: “All known oomycetes” implies you checked every oomycete ever described*
 
@@
 
-> 214: Which cutoffs did you use for the parameters such as GC content, hairpins etc when selecting primers?
+*214: Which cutoffs did you use for the parameters such as GC content, hairpins etc when selecting primers?*
 
-Rigid cutoffs were not used, but rather people experienced with primer design compared the OligoAnalyzer results to their previous experiences.
+Rigid cutoffs were not used, but rather people experienced with primer design compared the OligoAnalyzer results to their previous experiences to select primers to try in the lab.
 
-> 239: How was the DNA quantified?
+!!!!CHECK WITH NIK!!!!
+
+*239: How was the DNA quantified?*
 
@@
 
-> 321 – 322: This first sentence is not required because the information is repeated in more detail below. Please just add the references for DADA2 and Cutadapt where approcpriate below instead.
+*321 – 322: This first sentence is not required because the information is repeated in more detail below. Please just add the references for DADA2 and Cutadapt where approcpriate below instead.*
 
@@
 
-> 478 – 494: Were there any oomycetes amplified/identified with the ITS1 primer set that were not identified with the rps10 primers?
+*478 – 494: Were there any oomycetes amplified/identified with the ITS1 primer set that were not identified with the rps10 primers?*
 
@@
 
-* The content of the reference databases differs for the two loci, so even if both methods amplified the same organism, the closest match in the reference database of each locus could be to different taxa.
+1.
+The content of the reference databases differs for the two loci (about 1/3 of the species in each database are not in the other), so even if both methods amplified the same organism, the closest match in the reference database of each locus could be to different taxon.
 Since many ASVs in the environmental samples have no close match in either database (only 16% of oomycete ASVs from environmental samples had a match >95% identical), this error is likely to occur often.
-* The limited taxonomic resolution of ITS1, and rps10 to a lesser extent, means that even if both databases have a reference sequence perfectly matching an ASV, there could be other taxa in that database with the same sequence, so which taxon name gets assigned to the ASV is random.
+1.
+The limited taxonomic resolution of ITS1, and rps10 to a lesser extent, means that even if both databases have a reference sequence perfectly matching an ASV, there could be other taxa in that database with the same sequence, so which taxon name gets assigned to the ASV is random.
 
@@
 
-> 516 Please change “are generally more diverged from each other” to “are generally more divergent”
+!!!!CHECK WITH NIK!!!!
+
+*516 Please change “are generally more diverged from each other” to “are generally more divergent”*
 
@@
 
-> 540 - 541: Please change “showed” to “predicted” also please change “non-target taxa” to “all non-target taxa tested”
+*540 - 541: Please change “showed” to “predicted” also please change “non-target taxa” to “all non-target taxa tested”*
 
@@
 
-> This study describes the design of a valuable new barcoding locus/assay developed from a database of mtDNA sequence data.
+*This study describes the design of a valuable new barcoding locus/assay developed from a database of mtDNA sequence data.
 Its utility was then compared carefully against an ITS rDNA locus.
@@
 I also feel it may be helpful to detail which of the reverse primers has specificity to which group of oomycetes.
-Could this be used to give future researchers an option to pick and choose the combination of rps10 primers to focus on the group of oomycetes of particular interest?
+Could this be used to give future researchers an option to pick and choose the combination of rps10 primers to focus on the group of oomycetes of particular interest?*
 
-> As an aside: note that the primers ITS6 and 7 were never designed to be, or described as, an oomycete-specific assay by Cooke et al (2000) but were simply modifications of the existing White et al (1990) fungal ITS primers to allow more robust amplification of oomycete samples in the laboratory.
+!!!!TODO!!!!
+
+!!!!CHECK WITH NIK!!!!
+
+*As an aside: note that the primers ITS6 and 7 were never designed to be, or described as, an oomycete-specific assay by Cooke et al (2000) but were simply modifications of the existing White et al (1990) fungal ITS primers to allow more robust amplification of oomycete samples in the laboratory.
 They were taken up and tested by other authors as ‘oomycete specific’.
@@
 While a comparison may have been with the Scibetta et al (2012) ITS primers which are highly specific to known downy mildew and Phytophthora I understand that these do not amplify other oomycetes and would not have been equivalent.
-The key advantages of rps10 – oomycete specificity and improved taxonomic resolution - are clear.
+The key advantages of rps10 – oomycete specificity and improved taxonomic resolution - are clear.*
 
 These are excellent points.
-We have added a sentence to explain the history of ITS6 and ITS7.
+We have added a sentence to better explain the history of ITS6 and ITS7.
 
-> What happens with Nothophytophthora? I could not see them on Figure 2. There is no reason to suggest they would not amplify but perhaps they should be mentioned somewhere?
+*What happens with Nothophytophthora? I could not see them on Figure 2. There is no reason to suggest they would not amplify but perhaps they should be mentioned somewhere?*
 
@@
 At the time Nothophytophthora had just been described and we were not aware of any publicly available rps10 sequences for Nothophytophthora that included the primer binding sites.
-As far as we can tell, the rps10 sequences for Nothophytophthora on NCBI currently available do not have the primer binding sites so we cannot predict whether or not Nothophytophthora would be amplified by this method.
-Nothophytophthora is one of many genera not represented by currently available rsp10 sequences.
-We have added a sentence to the methods (around line 228) using Nothophytophthora as an example for genera we dont have data for.
+I just looked again and, as far as we can tell, the rps10 sequences for Nothophytophthora on NCBI currently available do not have the primer binding sites either so we cannot predict whether or not Nothophytophthora would be amplified by this method.
+Nothophytophthora is one of many genera not represented by currently available rps10 reference sequences.
+We have added a sentence to the methods (around line 228) using Nothophytophthora as an example for genera we don't have data for.
 
-> The web-site https://nam04.safelinks.protection.outlook.com/?url=http%3A%2F%2Fwww.oomycetedb.org%2F&amp;data=04%7C01%7Cniklaus.grunwald%40oregonstate.edu%7C04b24f72622c4cf9996308da05c111ce%7Cce6d05e13c5e4d6287a84c4a2713c113%7C0%7C0%7C637828624036753211%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000&amp;sdata=%2FlQvEfQI7iNfh7VRxfzITmsi4h2dliKEmNs2dGUVj80%3D&amp;reserved=0 is a valuable resource.
+!!!!CHECK WITH NIK!!!!
+
+*The web-site https://nam04.safelinks.protection.outlook.com/?url=http%3A%2F%2Fwww.oomycetedb.org%2F&amp;data=04%7C01%7Cniklaus.grunwald%40oregonstate.edu%7C04b24f72622c4cf9996308da05c111ce%7Cce6d05e13c5e4d6287a84c4a2713c113%7C0%7C0%7C637828624036753211%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000&amp;sdata=%2FlQvEfQI7iNfh7VRxfzITmsi4h2dliKEmNs2dGUVj80%3D&amp;reserved=0 is a valuable resource.*
 
@@
 
-> L145 “due to it robust statistical” needs an extra word or ‘its’
+*L145 “due to it robust statistical” needs an extra word or ‘its’*
 
@@
 
-> L153 “currently available ITS1” I would say “some currently available ITS1” as there are ones that do not such as the unmentioned Scibetta et al primers.
-I feel reference should be made to the source paper (Scibetta et al 2012) which described the downy mildew and Phytophthora-specific primers employed in several key studies such as those of Landa and Riddell.
-> Scibetta, S.; Schena, L.; Chimento, A.; Cacciola, S.O.; Cooke, D.E.
+*L153 “currently available ITS1” I would say “some currently available ITS1” as there are ones that do not such as the unmentioned Scibetta et al primers.
+I feel reference should be made to the source paper (Scibetta et al 2012) which described the downy mildew and Phytophthora-specific primers employed in several key studies such as those of Landa and Riddell.*
+*Scibetta, S.; Schena, L.; Chimento, A.; Cacciola, S.O.; Cooke, D.E.
 A molecular method to assess Phytophthora diversity in environmental samples.
@@
 Microbiol.
-Methods 2012, 88, 356–368, doi:10.1016/j.mimet.2011.12.012.
+Methods 2012, 88, 356–368, doi:10.1016/j.mimet.2011.12.012.*
 
@@
 
-> L157 “as little as 5.3% of” simpler just to say “only 5.3% of ”
+*L157 “as little as 5.3% of” simpler just to say “only 5.3% of ”*
 
@@
 
-> L185 suggest “power” or “specificity“ rather than ”usefulness” or leave and just say they were compared.
+*L185 suggest “power” or “specificity“ rather than ”usefulness” or leave and just say they were compared.*
 
@@
 
-> L294 should specify the version of Illumina chemistry employed.  v2 or v3 and how many cycles?  This is important.
+*L294 should specify the version of Illumina chemistry employed.  v2 or v3 and how many cycles?  This is important.*
 
@@
 
-> L471 onwards – for the mock community it would also be interesting to know the read numbers for each species.
+*L471 onwards – for the mock community it would also be interesting to know the read numbers for each species.
 Were all reads balanced in 1/24 ratio?
@@
 Was this the case?
-I understand the assays are not intended to be quantitative but it would be useful to know if differences in read number broadly align with initial DNA concentration.
+I understand the assays are not intended to be quantitative but it would be useful to know if differences in read number broadly align with initial DNA concentration.*
 
-Good idea. This information has been added.
+Good idea. This information has been added (around line 490).
 
-> SE Comment: Is there a spot you could include the importance to Phytobiome research, maybe early in the introduction. It is just nice to see the word phytobiome/s used.
+!!!!CHECK WITH NIK!!!!
+
+*SE Comment: Is there a spot you could include the importance to Phytobiome research, maybe early in the introduction. It is just nice to see the word phytobiome/s used.*
 

### 2022-04-01 publication/reviewer_response.Rmd
add multiple sequences alignment of amplicons
simplified primer data tables

@@
 
-!!!!CHECK WITH NIK!!!!
-
 *176: Change “amplification” to “PCR”*
@@
 
-!!!!TODO!!!! (Waiting for Frank to get back to us with the alignment)
+We have added an alignment of the region amplified (including the primer binding sites) derived from the reference database, which includes most of the sequences used in the initial alignment as well as many more oomycete sequences we did not have when designing the primers.
 
@@
 
-!!!!CHECK WITH NIK!!!!
-
 *239: How was the DNA quantified?*
@@
 
-*321 – 322: This first sentence is not required because the information is repeated in more detail below. Please just add the references for DADA2 and Cutadapt where approcpriate below instead.*
+*321 – 322: This first sentence is not required because the information is repeated in more detail below. Please just add the references for DADA2 and Cutadapt where appropriate below instead.*
 
@@
 
-!!!!CHECK WITH NIK!!!!
-
 *516 Please change “are generally more diverged from each other” to “are generally more divergent”*
@@
 
-Thanks for pointing that out. Done.
+Done.
 
@@
 
-!!!!TODO!!!!
-
-!!!!CHECK WITH NIK!!!!
+Thank you for your review and constructive feedback!
+We have added two supplementary tables that inform which primer variations correspond to each genus and species.
 
@@
 We have added a sentence to the methods (around line 228) using Nothophytophthora as an example for genera we don't have data for.
-
-!!!!CHECK WITH NIK!!!!
 
@@
 
-Good idea. This information has been added (around line 490).
-
-!!!!CHECK WITH NIK!!!!
+Good idea. This information has been added (around line 490 and in a supplemental figure).
 

### 2022-04-01 09--simulated_pcr_results.Rmd
add multiple sequences alignment of amplicons
simplified primer data tables

@@
 ```{r}
-write_csv(binding_data, file = file.path('results', 'primer_binding_site_sequences.csv'))
+binding_data %>%
+  select(species, forward, reverse, amplicon) %>%
+  write_csv(file = file.path('results', 'primer_binding_site_sequences.csv'))
+```
+
+An alignment of the amplicons might also be useful:
+
+```{r}
+aligned <- binding_data$amplicon %>%
+  insect::char2dna() %>%
+  ips::mafft(method = 'localpair', exec = '/usr/bin/mafft')
+write.FASTA(aligned, file = file.path('results', 'amplicon_region_multiple_alignment.fa'))
 ```
@@
 rownames(genus_bindings_data) <- NULL
-write_csv(as.data.frame(genus_bindings_data),
-          file = file.path('results', 'genus_primer_binding_site_counts.csv'))
+genus_bindings_data %>%
+  as_tibble() %>%
+  filter(type == "pair") %>%
+  select(-type) %>%
+  write_csv(file = file.path('results', 'genus_primer_binding_site_counts.csv'))
 ```
